---
title: "Covariate balance after matching"
subtitle: "Why within-set balance tests collapse, and what to report instead"
author: "Jake Bowers (with Claude)"
date: today
format:
  pdf:
    toc: true
    toc-depth: 3
    number-sections: false
    fontsize: 12pt
    geometry:
      - left=1.25in
      - right=1.25in
      - top=1in
      - bottom=1in
    fig-width: 6.5
    fig-height: 4
---

```{r}
#| label: setup
#| message: false
#| warning: false
#| echo: false
#| include: false
devtools::load_all(here::here(), quiet = TRUE)
options(scipen=8,digits=4)
```

This memo is about a problem in the Hansen-Bowers (2008) balance test, and about
what to report in its place. The problem: as a matched design is made tighter, the
within-set balance test from `balanceTest()` gives the design *worse* p-values,
even when the imbalance that remains is substantively trivial. A design that matched
better looks worse.

The memo is organized around the problem rather than around any one fix. It first
shows the collapse plainly --- in one dimension, where the arithmetic is a single
fraction, and then in many. It then argues that no change of statistic escapes the
collapse, and states the choice this forces as a trilemma. It recommends what to
report instead: an imbalance *magnitude* on a scale fixed before matching, together
with the per-covariate, substantive judgment a matcher already makes. It relates the
approach to Branson (2021), the closest prior work. Finally it discusses --- but does
not propose as the primary tool --- a resolution-profile summary, $l^*$, and a
separate high-dimensional pathology of the omnibus.

Where a claim is proved, shown only by simulation, or argued heuristically, I say so.

---

## The problem we are solving

We have a matched design. Units are grouped into sets. Each set holds treated and
control units (or, in a dose study, higher- and lower-dose units) that are
similar on the measured covariates. We want to argue that this design supports
the assumption behind the analysis: that within sets, treatment is as good as
randomly assigned. The evidence we have is the set of covariate differences
between treated and control units, summed across sets with some weights. We want a single summary
of those differences across all covariates at once, and we want to read that
summary against a standard --- what a randomized experiment would produce.

A randomized experiment is the right standard because it is the design we trust.
The question is which randomized experiment. Complete randomization of the whole
sample is one candidate, but it is the wrong one here: matching reorganizes the
sample into similar sets rather than discarding units, so the experiment it
imitates is one that randomizes treatment within blocks, not one that randomizes
the whole sample at once. Comparing instead against complete randomization asks
how much the matching improved on no matching at all --- the question behind
matching-as-pruning, where units are dropped until the groups balance and the
result is analyzed as if completely randomized. That is a different workflow from
the one here, where the matched sets are built and then held fixed, and we do not
take it as our standard.

---

## Why within-set balance tests collapse

The $d^2$ test compares the covariate imbalance we observe to the natural,
randomization-based, imbalance we would see if treatment were re-randomized
within the realized matched sets. As the match gets tighter a small, real,
persistent difference between treated and control units gets driven toward a
$p$-value of zero, so a design that matched better looks worse.

I don't think that this problem can be fixed by changing the statistic. Any statistic
computed from the matched sample, compared to a reference distribution built from
that same sample, returns the same verdict whether the within-set differences are
large or tiny --- as the worked alternatives at the end of this section show, and as
the appendix on metric-invariance argues formally.

Here is the mechanism in one line of algebra. Write $d$ for the vector of
treated-minus-control covariate differences summed across sets, and $V$ for the
covariance of $d$ under re-randomization within the matched sets. The $d^2$ statistic
is

$$ T = d' V^{+} d , $$

where $V^{+}$ is the Moore-Penrose pseudoinverse. It equals the ordinary inverse
$V^{-1}$ only when $V$ is full rank; $V$ is often rank-deficient here (more
covariate directions than the within-set re-randomization can move), which is why
`balanceTest()` forms $V^{+}$ by a truncated SVD and reports degrees of freedom
equal to the rank of $V$ rather than to the number of covariates. For the scaling
argument that follows the distinction does not matter --- the same cancellation
holds for $V^{+}$ --- but it is the entire story behind the separate
high-dimensional degeneracy of this test, discussed in its own section below, where
the rank of $V$ falls short of the covariate count and the omnibus loses its ability
to discriminate.

Now imagine shrinking every within-set distance by a factor $\delta$ (a tighter
match). The differences d shrink in proportion to $\delta$. The covariance $V$ is
built from within-set sums of squares, so it shrinks in proportion to $\delta^2$.
The ratio $T = d' V^{+} d$ then has $\delta$ in the numerator and $\delta^2$ in the
denominator, twice over, so the two cancel and $T$ does not change at all. The test
literally cannot tell a match with $\delta = 1$ from a match with $\delta = 0.001$.

This is not special to $d^2$, and it is not special to quadratic forms. Three
standard alternatives mark the range. Regress treatment on the covariates and read
off the joint F-test; compare nested logistic regressions of treatment on the
covariates by a likelihood-ratio test; or, refusing the quadratic form altogether,
take the largest of the per-covariate standardized within-stratum differences and read
its p-value off a permutation reference --- the Westfall-Young max-$T$ omnibus, here
through `coin::independence_test`. The first two are the quadratic form's
least-squares and likelihood relatives; the third is a different construction --- a
maximum of marginals, not a sum of squares. All of them standardize each covariate's
difference by the same within-stratum spread that matching shrinks, so all of them
collapse. On one small design, as the within-set noise shrinks while a fixed gap on
$x_1$ stays put:

```{r}
#| label: competition
#| warning: false
## a matched design with a fixed gap on x1 and a tunable within-set noise scale
make_sets <- function(shrink, nsets = 6, K = 5, gap = 0.4, seed = 99) {
  set.seed(seed); rows <- list(); z <- blk <- numeric(0)
  for (s in seq_len(nsets)) {
    cen <- rnorm(K, sd = 5); g <- c(gap, rep(0, K - 1))
    rows[[s]] <- rbind(cen + g/2 + shrink * rnorm(K), cen + g/2 + shrink * rnorm(K),
                       cen - g/2 + shrink * rnorm(K), cen - g/2 + shrink * rnorm(K))
    z <- c(z, 1, 1, 0, 0); blk <- c(blk, rep(s, 4))
  }
  X <- do.call(rbind, rows); colnames(X) <- paste0("x", seq_len(K))
  data.frame(X, z = z, mset = factor(blk))
}

compete <- sapply(c(0.5, 0.2, 0.05, 0.01), function(sh) {
  d <- make_sets(sh)
  ## (1) HB08 d^2, within-stratum permutation covariance
  p_d2 <- balanceTest(z ~ x1 + x2 + x3 + x4 + x5 + strata(mset), data = d)$overall["mset", "p.value"]
  ## (2) F-test: treatment on covariates with stratum fixed effects, joint F
  p_F  <- anova(lm(z ~ mset, data = d),
                lm(z ~ x1 + x2 + x3 + x4 + x5 + mset, data = d))[["Pr(>F)"]][2]
  ## (3) logistic-regression likelihood-ratio test of the covariates
  gr <- glm(z ~ mset, family = binomial, data = d)
  gf <- glm(z ~ x1 + x2 + x3 + x4 + x5 + mset, family = binomial, data = d)
  p_LR <- anova(gr, gf, test = "LRT")[["Pr(>Chi)"]][2]
  p_coin_maxT <- coin::pvalue(coin:::independence_test(x1 + x2 + x3 + x4 + x5~z|factor(mset),data=d,teststat="maximum",distribution="asymptotic"))
  c(shrink = sh, d2_p = p_d2, Ftest_p = p_F, logitLR_p = p_LR, coin_maxT=p_coin_maxT)
})
signif(t(compete), 3)
```

At the loose match the omnibus tests do not reject --- the $d^2$, F, and max-$T$
p-values all sit near $0.2$ --- yet as the match tightens all four fall toward zero,
though the imbalance on $x_1$ never changed. The $d^2$ test flattens near $0.003$: its
within-stratum reference is discrete and finitely supported, and no observed
assignment can be more extreme than the most extreme reshuffle, which puts a floor
under its p-value. The F and max-$T$ tests keep descending, their continuous
large-sample references imposing no such floor; the likelihood-ratio p-value collapses
as well, flattening only because the fitted logistic regression separates.

The max-$T$ result is the one to dwell on. It is not a quadratic form, so it escapes
the separate high-dimensional degeneracy discussed later --- the rank ceiling that
freezes $d^2$ once the covariates outnumber the within-block residual degrees of
freedom. It is in fact the same kind of marginal-combination omnibus that the
high-dimensional section turns to for that case (the Cauchy combination there is its
close relative). And it collapses here exactly as the quadratic forms do. Escaping
the rank ceiling does not escape the collapse, because the collapse lives in the
reference, not in how the per-covariate signals are pooled: each marginal
standardized difference is itself measured against the within-set spread, so each is
driven toward rejection as the match tightens. The appendix on metric-invariance
gives the formal reason no fixed change of statistic --- quadratic, likelihood, or
maximum --- can help under the within-set reference.

## A worked example in one dimension

With one covariate the covariance $V$ is a scalar. The $d^2$ statistic reduces to a plain fraction,

$$
T = \frac{d^2}{V},
$$

which is just the square of the familiar standardized difference $d/\sqrt{V}$
that `balanceTest()` already prints for each covariate. Everything in the collapse
argument above is visible in that one fraction.

Write the pieces out for a single covariate $x$, a binary treatment $z$, and
matched sets $b$. Let $n_b$, $n_{1b}$, $n_{0b}$ be the set's total, treated, and
control counts, and let $\bar x_{1b}, \bar x_{0b}$ be the treated and control
means within the set. Then the summed adjusted difference and its
within-set re-randomization variance are

$$
d = \sum_b \frac{n_{1b}\,n_{0b}}{n_b}\,(\bar x_{1b} - \bar x_{0b}),
\qquad
V = \sum_b \frac{n_{1b}\,n_{0b}}{n_b\,(n_b-1)}\;
        \sum_{i \in b} (x_i - \bar x_b)^2 .
$$
The first quantity, $d$, is the imbalance the design shows: the treated-minus-
control difference on $x$, summed across sets. The second is its yardstick. The
inner sum $\sum_{i \in b}(x_i - \bar x_b)^2$ is the spread of $x$ within set $b$
--- the sum of squared distances of the set's units from their own mean --- and
the leading factor $n_{1b}\,n_{0b} / \{n_b(n_b-1)\}$ turns that spread into the
variance of $d$ produced by reshuffling which units in each set are treated. So
$V$ is not itself an imbalance; it is how much $d$ would wobble under within-set
re-randomization, and $T = d^2/V$ weighs the observed imbalance against that
wobble. The two helpers below compute exactly $d$ and $V$ --- the one-covariate
special case of `randomization_cov_d` in `balanceTestEngine`.

```{r}
#| label: oned-helpers
## d = sum_i (z_i - pi_b) x_i, with pi_b the within-set treated fraction
adj_diff <- function(x, blk, z) sum((z - ave(z, blk, FUN = mean)) * x)

## V = within-set re-randomization variance of d (scalar, one covariate)
perm_var <- function(x, blk, z) {
  v <- 0
  for (idx in split(seq_along(x), blk)) {
    n <- length(idx); n1 <- sum(z[idx]); n0 <- n - n1
    if (n1 == 0 || n0 == 0 || n < 2) next
    v <- v + (n1 * n0 / (n * (n - 1))) * sum((x[idx] - mean(x[idx]))^2)
  }
  v
}

## $d^2$ statistic for one covariate: a plain ratio, no matrix inverse
d2_1d <- function(x, blk, z) adj_diff(x, blk, z)^2 / perm_var(x, blk, z)
```

Now a small design to compute on: six matched sets, each with two treated and
two control units. Treated units sit a fixed gap above their controls on $x$;
the parameter `shrink` scales the within-set noise, so a smaller `shrink` is a
tighter match.

```{r}
#| label: oned-data
make_1d <- function(shrink, gap = 0.5, nsets = 6) {
  set.seed(5)
  x <- z <- blk <- numeric(0)
  for (s in seq_len(nsets)) {
    ctr <- 10 * s
    x   <- c(x, ctr + gap / 2 + shrink * rnorm(2),   # two treated
                ctr - gap / 2 + shrink * rnorm(2))   # two controls
    z   <- c(z, 1, 1, 0, 0)
    blk <- c(blk, rep(s, 4))
  }
  data.frame(x = x, z = z, mset = factor(blk))
}
head(make_1d(shrink = 0.3), 8)
```

The hand-computed ratio is exactly what `balanceTest()` reports as the omnibus
chi-square for this one-covariate design. There is no covariance to worry about,
so the two agree to machine precision.

```{r}
#| label: oned-byhand
d0 <- make_1d(shrink = 0.3)
dval <- adj_diff(d0$x, d0$mset, d0$z)
Vval <- perm_var(d0$x, d0$mset, d0$z)
bt   <- balanceTest(z ~ x + strata(mset), data = d0)
round(c(d = dval, V = Vval,
        T_byhand      = dval^2 / Vval,
        T_balanceTest = bt$overall["mset", "chisquare"]), 5)
```

### Reading 1: the test is blind to the scale of the match

The claim above is that shrinking every within-set distance by a factor
$\delta$ leaves $T$ unchanged, because $d$ scales like $\delta$ and $V$ scales
like $\delta^2$. Take the design above, hold its shape fixed, and scale every
unit's deviation from its set center by $\delta$. The imbalance $d$ and the
variance $V$ both shrink --- but at different rates, and the ratio does not move.

```{r}
#| label: oned-invariance
base <- make_1d(shrink = 0.3)
ctr  <- ave(base$x, base$mset)   # set centers
dev  <- base$x - ctr             # within-set deviations
invariance <- sapply(c(1, 0.1, 0.01, 0.001), function(delta) {
  x <- ctr + delta * dev
  c(delta = delta,
    d = adj_diff(x, base$mset, base$z),
    V = perm_var(x, base$mset, base$z),
    T = d2_1d(x, base$mset, base$z))
})
round(t(invariance), 6)
```

The `d` column falls tenfold at each step, the `V` column a hundredfold, and the
`T` column does not move. A match a thousand times tighter gets the identical
statistic. The reason is that $T = d^2/V$ is a ratio --- the imbalance measured in
units of the within-set spread --- so scaling the whole configuration by $\delta$
shrinks the imbalance and the spread together and leaves the ratio fixed. The test
therefore reads only the imbalance relative to the spread of the very covariates
the matching compressed; it never sees the imbalance on its own, natural scale. A
design with a half-unit imbalance against a unit of spread and a design with a
thousandth-unit imbalance against a thousandth of spread look identical to it, even
though the second is matched a thousand times more tightly.

### Reading 2: a fixed residual makes a better match look worse

The scaling above moved the gap and the noise together. The situation an analyst
actually faces is different: matching shrinks the within-set noise but leaves a
small, persistent gap it could not remove. Hold the gap fixed and shrink only the
noise. Now $d$ stays put while $V$ falls toward the variance the gap alone
implies, so $T$ rises and the p-value falls --- the design that matched its noise
away more tightly is flagged harder.

```{r}
#| label: oned-collapse
collapse <- sapply(c(0.5, 0.2, 0.05, 0.01), function(sh) {
  d0 <- make_1d(shrink = sh)
  Tn <- d2_1d(d0$x, d0$mset, d0$z)
  c(shrink = sh, T = Tn, p = pchisq(Tn, df = 1, lower.tail = FALSE))
})
round(t(collapse), 5)
```

The p-value drops from about 0.02 to about 0.00002 as the match tightens, even
though the real treated-minus-control gap never changed. It plateaus near a
chi-square of 18 rather than running all the way to zero, because here the fixed
gap also contributes to the within-set scatter and so puts a floor under $V$;
with more covariates and coarser within-set noise the descent is steeper, as the
next example shows. Either way the direction is the wrong one: tightening the
match worsens the verdict.

### From one covariate to many: the same collapse through `balanceTest()`

The one-covariate case made the arithmetic transparent. Nothing about the
collapse depended on having a single covariate, though. Here is the omnibus test
through `balanceTest()` on a more realistic design: forty matched sets of four,
$N = 160$, five covariates, with a fixed treated-minus-control gap on the first
covariate and the rest balanced in expectation.

```{r}
#| label: mv-data
make_mv <- function(shrink, nsets = 40, K = 5, gap = 0.4, seed = 99) {
  set.seed(seed)
  rows <- list(); zz <- bb <- numeric(0)
  for (s in seq_len(nsets)) {
    cen <- rnorm(K, sd = 5) + 0.1 * s     # set center drifts across sets
    g   <- c(gap, rep(0, K - 1))          # persistent gap on x1 only
    rows[[s]] <- rbind(cen + g / 2 + shrink * rnorm(K),
                       cen + g / 2 + shrink * rnorm(K),
                       cen - g / 2 + shrink * rnorm(K),
                       cen - g / 2 + shrink * rnorm(K))
    zz <- c(zz, 1, 1, 0, 0); bb <- c(bb, rep(s, 4))
  }
  X <- do.call(rbind, rows); colnames(X) <- paste0("x", seq_len(K))
  data.frame(X, z = zz, mset = factor(bb))
}

mv <- sapply(c(0.8, 0.3, 0.1, 0.03), function(sh) {
  bt <- balanceTest(z ~ x1 + x2 + x3 + x4 + x5 + strata(mset),
                    data = make_mv(sh))
  o  <- bt$overall["mset", ]   # one row of the omnibus table
  c(shrink = sh, chisquare = o[["chisquare"]],
    df = o[["df"]], p.value = o[["p.value"]])
})
signif(t(mv), 4)   # signif, not round, so the tiny p-values survive
```

The omnibus chi-square climbs from about 20 to about 118 and the p-value falls
from roughly $10^{-3}$ to roughly $10^{-23}$ as the match tightens, on five
degrees of freedom throughout. The persistent gap on $x_1$ never changed. The
same fraction is at work as in one dimension: the residual stays while the
within-set noise that anchors $V$ disappears, so the test reads a tighter,
better match as stronger evidence against the design.

---

## What the collapse means: a trilemma

The collapse is not a defect of $d^2$ in particular. The competing-statistics
table and the appendix on metric-invariance say that any statistic built from the
matched sample, compared to a reference drawn from that same sample, collapses the
same way. This is worth stating as a constraint, because it explains why the
problem cannot be fixed cheaply and what the real choices are.

The standard a matched analysis actually assumes is re-randomization of treatment
*within the realized matched sets* --- the design the matching was built to imitate.
Against that standard, any calibration we compute (a p-value, a percentile, "more
balanced than X% of randomized versions of this design") collapses as the match
tightens, because the reference distribution's spread is built from the within-set
covariate spread, and tightening the match is the act of shrinking that spread. The
observed imbalance then sits in an ever-narrower reference, and the verdict turns
against the design regardless of which statistic or which fixed metric we choose.

So one cannot have all three of these at once:

1. the within-set standard (re-randomize treatment within the realized sets);
2. a calibration --- "better than X% of randomized versions" --- that does not
   collapse as the match tightens;
3. no comparison to designs the analyst did not build.

Pick two. Reading the imbalance as a *magnitude* on a fixed external scale, rather
than as a calibration, keeps (1) and (3) and gives up (2): it reports how large the
imbalance is, not how surprising it is under a shrinking reference. Comparing the
design against coarser reference blockings keeps (2) and gives up part of (3): it
buys a non-collapsing calibration by relaxing the within-set standard to a looser
one. The resolution-profile section below is that second corner; the recommendation
that follows is the first.

One point the collapse settles, and that matters for reading the within-set
p-value at all: a small p-value there is often *correct detection*. A persistent,
same-sign covariate gap across sets genuinely is not what within-set randomization
produces, and the test is right to flag it. What the p-value cannot do is say
whether the gap it flags is substantively large or trivially small --- because its
verdict reflects the precision of the match, not the size of the imbalance. That
gap between significance and magnitude is what the next section addresses.

---

## What to report instead

The recommendation is to report the imbalance as a magnitude on a scale fixed
before matching, and to keep the per-covariate, substantive judgment a matcher
already makes, rather than to read a single collapsing p-value as a measure of
design quality.

**A magnitude on a fixed scale.** Replace the within-set covariance $V$, which
shrinks with the match, by a covariance fixed before matching --- the covariance of
the covariates in the pre-match pool, $\Sigma_{\text{pool}}$ --- and report the
Mahalanobis imbalance $d'\,\Sigma_{\text{pool}}^{-1}\,d$ as a number, not as a
p-value. Because the denominator no longer moves with the match, this magnitude
falls as the match improves: it sees the imbalance on a scale the matching did not
compress. The appendix on metric-invariance shows the contrast directly --- the
within-set *p-value* is the same whatever fixed metric is used, but the *magnitude*
the fixed metric produces is precision-invariant and shrinks as the noise shrinks.
This omnibus magnitude is the same statistic Branson (2021) uses (the next section),
read as a magnitude rather than as a design-selection p-value.

**Per-covariate differences and substantive judgment.** The omnibus magnitude
summarizes; the matcher's actual craft is per-covariate. For the covariates that
matter, report the treated-minus-control differences within sets and their
distribution across sets --- on a scale fixed before matching --- and exercise
judgment on them with collaborators: "half the sets show no age difference, but two
sets differ by more than ten years; is that acceptable here?" `balanceTest()`
already prints per-covariate standardized differences; the within-set-difference
percentiles across sets are the part worth adding. This is not a softening of the
analysis. It is the recognition that whether a residual imbalance matters is a
substantive question about the covariate and the outcome, not one a single number
can settle.

**An equivalence test, if a calibrated statement is wanted.** A matcher who wants a
yes/no calibration that does not collapse can test equivalence rather than
difference: instead of testing "is there any imbalance," test "is the imbalance
smaller than a stated tolerance $\delta$." This keeps the within-set standard,
invokes no other design, and behaves correctly as the match tightens --- a tighter
match makes equivalence easier to conclude. The cost is that the tolerance $\delta$
must be named, which is the same "how large is too large" judgment as above, now
made explicit on the covariate scale. A design-respecting equivalence procedure
needs a proper null for the fixed-metric magnitude; that is the `sigma_x`
machinery on the development branch, and it is the natural next implementation step
if the team wants the calibrated form.

---

## Relation to Branson (2021)

Consider what an analyst actually does. She has a pool of treated and control units,
she matches them into sets that are similar on the measured covariates, and she
builds those sets to approximate a block-randomized experiment --- in the simplest
case, a paired experiment in which one unit of each pair would have been assigned to
treatment by the flip of a coin. She then holds that matched design fixed and
analyzes the outcomes as the design she built, conditioning on the sets she formed.
The covariate balance she reports is evidence for the assumption her analysis rests
on: that within sets, treatment is as good as randomly assigned. She has already
specified the experiment her matching imitates, so her question is not which
experiment her data resemble, but whether the match she built supports that
specification well enough on the covariates she can see.

The work we build on here is Branson (2021), who formalized the test of whether a
matched dataset is consistent with a specified design's randomization. Testing
balance against a design-induced reference is already what Hansen and Bowers (2008)
do, conditioning on the within-block design; Branson's sharpening is to make the
design an explicit, swappable null. His $H_0$ is that treatment was drawn from
$P^*(W \mid X)$ for a particular $P^*$, and he re-runs the same test under each
candidate $P^*$ in turn: complete randomization, block or paired randomization, and
constrained (rerandomized) assignment. For an omnibus summary he recommends a
Mahalanobis distance, $M = (\bar x_T - \bar x_C)' [\text{cov}(\bar x_T - \bar
x_C)]^{-1} (\bar x_T - \bar x_C)$, in which $\bar x_T - \bar x_C$ is the overall,
unblocked treated-minus-control mean difference and $\text{cov}(\bar x_T - \bar x_C)
= (N / N_T N_C)\,\text{cov}(X)$ is the full pre-match pool covariance of the
covariates, computed once and held fixed across every design. Because the metric is
fixed, the reference distributions of the several designs can be drawn on one scale
and overlaid, with a single line at the observed $M$; Branson reads which design a
dataset approximates by where that one value falls in each density --- preferring
the design under which it is most central, flagging a dataset as too well balanced
for one design (evidence for a more constrained one) or too imbalanced to be
plausible under another. He calls his test a generalization of Hansen and Bowers
(2008), with their permutation test a special case in which draws from $P^*$ are
permutations of the observed assignment. Making the design an explicit, swappable
null and putting several designs on one fixed scale are Branson's contributions, and
we use both.

One construction choice separates his test from the Hansen-Bowers $d^2$, and it is
two-fold. Branson's input is the unblocked overall mean difference and his
denominator is the fixed pool covariance; the Hansen-Bowers omnibus is
$T = d'\,V_d^{+}\,d$, where $d$ is the within-block-adjusted treated-minus-control
difference vector and $V_d$ is the covariance of $d$ under within-block permutation
--- a quantity that depends on the design and on the realized match, and that shrinks
as the matched sets tighten. So Branson holds one metric fixed and slides the design
into the reference, while the Hansen-Bowers test studentizes a within-block-adjusted
difference by a denominator that moves with the match. That moving denominator is
exactly what produces the collapse above, which Branson's fixed-metric,
single-resolution comparison does not exhibit. His $M$ is outside the collapse for
two reasons worth keeping separate: his statistic is the unblocked overall
difference, not a within-stratum-deviation quantity, and he compares designs at a
single blocking rather than refining it. His fixed metric is not what protects him;
an unblocked input and a single resolution are.

This is why our recommendation reports Branson's statistic as a magnitude. The
quantity $d'\,\Sigma_{\text{pool}}^{-1}\,d$ we report uses his pool covariance
(his footnote 3, p. 9, notes that `randChecks` computes $\text{cov}(X)$ on the full
unmatched pool by default), applied to the within-block-adjusted difference our
blocked design produces, and read as a magnitude rather than a design-selection
p-value. We decline the design-menu calibration, and one end of it in particular:
reading a paired design against complete randomization is the matching-as-pruning
question of the opening section, which we do not adopt. Branson's own caution is
consistent with holding the built design fixed: he warns (p. 19) that "it may be
harmful to assume a precise experimental design or even any design if there are
substantial biases that remain after matching."

*Citation to verify before circulating:* Zach Branson, "Randomization Tests to
Assess Covariate Balance When Designing and Analyzing Matched Datasets,"
Observational Studies 7(2), 2021, pp. 44-80 (page headers; the Project MUSE cover
says pp. 1-36). DOI 10.1353/obs.2021.0031; arXiv:1804.08760. `randChecks` on CRAN.
The quotes above come from a read of the PDF; spot-check them and the page numbers
before publication.

---

## An optional sensitivity analysis: the resolution profile

This section records an idea the team developed and then chose not to recommend as
the primary tool. It is the second corner of the trilemma --- buy a non-collapsing
calibration by relaxing the within-set standard to a coarser one --- and it is
included as an optional sensitivity analysis, not as the memo's proposal. A matcher
who holds the design fixed and reports a magnitude (the recommendation above) does
not need it.

The idea. Re-randomizing treatment within the realized matched sets is the finest
of a family of block-randomized references; coarser members merge nearby matched
sets into larger blocks, up to one block holding everyone (complete randomization).
At any blocking $B$, compute the design's percentile in the within-block
re-randomization distribution of the same $d^2$ statistic: the fraction of
within-block-shuffled assignments whose imbalance is at least as large as the one
the design shows. Sweep the percentile across the family and plot it; this is the
**resolution profile**. Its finest end reproduces the $d^2$ test, which collapses;
its coarsest end reproduces the complete-randomization comparison --- the
matching-as-pruning question we do not adopt as a standard, present here only as the
endpoint. The single summary is the crossing:

    l* = the finest reference blocking at which the design's percentile first
         reaches 1/2 --- the finest block-randomized experiment the design
         balances the measured covariates as well as.

The honest reading of $l^*$ is not "as good as a design you did not build" but a
sensitivity analysis on the *resolution* of the within-set exchangeability
assumption: a coarser block of nearby units is a weaker form of the same assumption
(exchangeable among roughly-similar units, not nearly-identical ones), and $l^*$
locates the strongest version of that assumption the data do not contradict. Its
coarse, complete-randomization end is the pruning question and is kept distinct from
its block-randomized middle.

### Why coarsening the reference rescues the calibration

The collapse came from the within-set covariance $V$ shrinking as fast as the
differences $d$. A coarser reference breaks that link. When the reference blocks are
coarser than the matched sets, each block contains units from several matched sets,
so the covariate spread within a block includes the differences between set centers.
Those between-set differences are fixed by where the sets sit; they do not shrink
when you tighten the match. The coarse reference's covariance stays bounded away from
zero as $\delta \to 0$, while the imbalance it measures shrinks toward the residual
the matching could not remove. The numerator shrinks, the denominator does not, and
the percentile rises as the match improves --- the right direction. Status: the
collapse at the finest reference is established; the immunity of coarser references
is a scaling argument plus simulation, not a general proof.

### Building the ladder, and two findings

The ladder must be pre-specified and geometry-respecting. A block in a real
block-randomized experiment holds units that resemble each other, so the only honest
coarsening merges the matched sets that are *nearest* on the matching score the
design was built from (a set located by, say, the average or median score of its
units), declared before balance is examined. This is not a free parameter:
merging far-apart sets instead flatters the design by a full rung. On one fixed
design, the geometry-respecting and interleaved ladders give $l^*$ of about 11.5 and
6.4 units per block --- larger than the bin-offset wiggle the stability study
reports, and always in the design-flattering direction. Pre-specifying the
nearest-set-merge rule closes this degree of freedom.

A second finding cuts the other way and answers the standard objection that balance
is about the covariate distributions, not about who is paired with whom. $l^*$ is
invariant to the pairing: holding the same units and the same treated and control
covariate distributions, three different valid pairings of the same units give
resolution profiles that agree to about 0.001. Above the finest rung the percentile
depends on the matched-set means and on which units sit in each resolution band, not
on who is paired with whom inside a set. Only the single finest rung uses the
pairing, and that is the rung that collapses.

The simulation behind these claims is in `vignettes/resolution-profile-demo.R`
(the collapse and the discrete ladder), `resolution-profile-stability.R` (the
continuous caliper, the $l^*$-vs-gap responsiveness, and the robustness study), and
`resolution-profile-threebenchmark.R` (the geometry-respecting ladder, the
pairing-invariance check, and a simulation-free rendering against a chance floor).
In brief: the collapse is confined to the single finest reference; $l^*$ rises
monotonically with the persistent gap; the analytic chi-square percentile matches a
within-block permutation except at the finest rung, which sits below $l^*$.

### How to read the uncertainty

The inference model is the assignment mechanism: conditional on the units, their
covariates, and the blocks, the only randomness is which units are treated. The
percentile is an exact, finite-sample position of the observed design in the
within-block re-randomization distribution, computed by enumeration when the support
is small or estimated by drawing assignments. There is no superpopulation and no
resampling, so there is no bootstrap. For one design the curve is the whole answer
and $l^*$ is one reading of it. The 30-to-40 percent spread of $l^*$ reported in the
stability study is the variation of $l^*$ across many designs of the same nominal
quality, not an uncertainty band for one analyst's one design.

### What is established, and what is not

Established: the $d^2$ collapse at the matched-set reference, and that no statistic
from the matched sample compared to a same-sample reference escapes it. Shown by
simulation and a scaling argument, not proved: that coarser references do not
collapse; that $l^*$ rises monotonically with the residual gap; that $l^*$ is
invariant to the pairing and stable once the ladder is geometry-respecting. Open: a
formal proof that coarser references escape the collapse; a unit for $l^*$ that is
comparable across set structures; and a prototype of the continuous caliper in
`balanceTest()`.

It speaks to the measured covariates only. A small $l^*$ says the design balances the
measured covariates as well as a finely blocked experiment would; it says nothing
about unmeasured confounders, and "as well as experiment $l^*$" is a statement of
consistency, not a certificate. Report it next to a sensitivity analysis for
unmeasured confounding, not in place of one.

---

## A second pathology: high dimensionality

The collapse is one way the omnibus fails. There is a second, unrelated one, and the
resolution profile addresses it only by accident and only halfway, so it is worth
keeping separate.

When the covariate count $p$ reaches or passes the within-block residual degrees
of freedom --- the total number of units $N$ minus the number of blocks $B$ ---
the statistic $T = d' V^{+} d$ stops responding to the covariates altogether. It
freezes at its own degrees of freedom, and the chi-square p-value freezes near
$0.46$.

The mechanism is exact, not asymptotic. Write the within-block-centered treatment
contrast $c_i = z_i - \pi_{b(i)}$ and let $X_c$ be the within-block-centered
covariate matrix, so that $d = X_c' c$ and $V = X_c'\,\mathrm{Cov}(c)\,X_c$, with
$\mathrm{Cov}(c)$ the within-block randomization covariance of the contrast. Then

$$
T = c'\,X_c \,\bigl(X_c'\,\mathrm{Cov}(c)\,X_c\bigr)^{+}\, X_c'\, c .
$$

When the centered covariates span the entire within-block residual space --- which
is exactly what $p \ge N - B$ buys --- the map $c \mapsto X_c' c$ is injective on
the contrast space and $T$ reduces to $c'\,\mathrm{Cov}(c)^{+} c$, the Mahalanobis
length of the contrast under its own covariance. That length is identical for every
assignment: all within-block assignments are permutations of one another and
$\mathrm{Cov}(c)$ is exchangeable, so the quadratic form is assignment-invariant and
equals its mean, the rank $N - B$. The covariates have cancelled out of the
statistic. $T = N - B$ is a constant, and the test reports
$P(\chi^2_{N-B} > N-B)$, which for any moderate rank sits near $0.46$.

```{r}
#| label: highdim-sweep
#| warning: false
## K strata of n_s units (n1 treated), p PURE-NOISE covariates (the null is true).
make_noise <- function(K, n_s, n1, p, seed = 2024) {
  set.seed(seed)
  z <- blk <- numeric(0); Xl <- list()
  for (s in seq_len(K)) {
    Xl[[s]] <- matrix(rnorm(n_s * p), n_s, p)
    zz <- rep(0, n_s); zz[sample(n_s, n1)] <- 1
    z <- c(z, zz); blk <- c(blk, rep(s, n_s))
  }
  X <- do.call(rbind, Xl); colnames(X) <- paste0("x", seq_len(p))
  data.frame(X, z = z, mset = factor(blk))
}

## fix the design (10 strata of 4, so residual df = N - B = 40 - 10 = 30); raise p
sweep_p <- sapply(c(2, 10, 20, 30, 50, 100), function(p) {
  d <- make_noise(K = 10, n_s = 4, n1 = 1, p = p)
  f <- reformulate(c(paste0("x", seq_len(p)), "strata(mset)"), response = "z")
  o <- balanceTest(f, data = d)$overall["mset", ]
  c(p = p, resid_df = 30, chisq = round(o[["chisquare"]], 1),
    used_df = o[["df"]], chisq_over_df = round(o[["chisquare"]] / o[["df"]], 3),
    p.value = round(o[["p.value"]], 3))
})
t(sweep_p)
```

Past $p = 30$ the statistic is pinned at $30$ and twenty more covariates change
nothing: the chi-square equals its degrees of freedom and the p-value sits at
$P(\chi^2_{30} > 30) = 0.466$. That pairing --- an omnibus p-value parked near
$0.45$ together with a chi-square almost exactly equal to its degrees of freedom
--- is the diagnostic to watch for. `balanceTest()` already warns when it happens
("Degrees of freedom exceeds units less number of strata...").

Does the resolution profile help? Coarsening the reference raises the residual
degrees of freedom $N - B$, so it can lift the design out of the saturated regime.
Below, $p = 30$ noise covariates with $N = 40$ units: the test is frozen at the
fine blockings and comes back to life once $N - B$ passes $30$.

```{r}
#| label: highdim-coarsen
#| warning: false
set.seed(13)
X <- matrix(rnorm(40 * 30), 40, 30); colnames(X) <- paste0("x", 1:30)
z <- rep(c(1, 0), 20)
coarsen <- sapply(c(20, 10, 5, 2, 1), function(B) {
  blk <- factor(rep(seq_len(B), each = 40 / B))
  f   <- reformulate(c(paste0("x", 1:30), "strata(mset)"), response = "z")
  o   <- balanceTest(f, data = data.frame(X, z = z, mset = blk))$overall["mset", ]
  c(blocks = B, resid_df = 40 - B, used_df = o[["df"]],
    chisq_over_df = round(o[["chisquare"]] / o[["df"]], 3),
    p.value = round(o[["p.value"]], 3))
})
t(coarsen)
```

The rescue is real but partial and impure. It works only when $p < N$: once
$p \ge N$ even complete randomization (rank $N-1$) is saturated, and no coarsening
can manufacture rank the sample does not hold. And the rescue arrives on the same
coarsening that loosens the reference, so it comes mixed with the leniency of the
resolution-profile section rather than as a clean correction.

The clean correction is a different statistic, not a different reference. The
pathology is that the quadratic form inverts a $p \times p$ covariance whose rank
is capped by the design. A test that never inverts that covariance --- one that
combines the per-covariate marginal tests, each of which is perfectly well behaved
--- has no rank ceiling and stays valid when $p$ exceeds $N$. The Cauchy
combination (ACAT) test of Liu and Xie (2020) is the natural choice: it combines
the per-covariate p-values `balanceTest()` already reports, is analytically robust
to dependence among the covariates, and is aimed exactly at the case where each
covariate admits a sensible one-dimensional statistic but the joint omnibus does
not. It senses sparse, marginal departures rather than the coordinated
multivariate shift the $d^2$ quadratic is tuned for, so the two are complements:
report ACAT beside $d^2$, not in place of it.

### The combination p-value, done correctly

The ACAT combination is one line of arithmetic. Given per-covariate p-values
$p_1, \dots, p_d$ (each uniform under the null) and weights $w_k$ summing to one,
form $T = \sum_k w_k \tan\{(0.5 - p_k)\pi\}$ and read off
$p_{\mathrm{ACAT}} = 0.5 - \arctan(T)/\pi$, the upper tail of a standard Cauchy.
The transform sends a small $p_k$ to a large positive term and a large $p_k$ to a
large negative one, and the Cauchy distribution's stability under averaging is what
makes the combination valid even when the covariates --- and hence their p-values
--- are dependent (Liu and Xie 2020).

```{r}
#| label: acat-corrected
acat_pvalue <- function(p, weights = NULL) {
  p <- p[!is.na(p)]
  d <- length(p)
  if (d == 0L) return(NA_real_)
  if (is.null(weights)) weights <- rep(1 / d, d) else weights <- weights / sum(weights)
  p <- pmin(p, 1 - .Machine$double.eps)        # tan -> -Inf exactly at p = 1
  small <- p < 1e-16                            # tan((0.5-p)pi) ~ 1/(p pi) for tiny p
  stat <- if (any(small)) {
    sum((weights[small] / p[small]) / pi) +
      sum(weights[!small] * tan((0.5 - p[!small]) * pi))
  } else {
    sum(weights * tan((0.5 - p) * pi))
  }
  0.5 - atan(stat) / pi
}
```

The test of a combination rule is its calibration: under the global null the
per-covariate p-values are uniform, so the combined p-value must be uniform too.

```{r}
#| label: acat-calibration
set.seed(2024)
B <- 4000; K <- 6
null_p <- replicate(B, acat_pvalue(runif(K)))  # K independent covariates, null true
round(c(mean = mean(null_p),                    # uniform -> 0.5
        type_I_at_0.05 = mean(null_p < 0.05),   #         -> 0.05
        type_I_at_0.10 = mean(null_p < 0.10)),  #         -> 0.10
      3)
```

The mean sits at $0.5$ and the rejection rates track their nominal levels. This is
worth stating because the prototype on the `cauchycomb` branch does not pass it:
that version halves each p-value before the transform and then doubles a minimum at
the end, two steps that are not in the published method and that break the Cauchy
calibration (its null mean is about $0.31$ and its rejection rate at $0.10$ is about
$0.14$). The function above --- feed the uniform p-values straight in, take the
standard Cauchy tail --- is the published rule, and it is the correction the branch
needs.

---

## Appendix A. The collapse does not depend on the statistic

The collapse section claims that comparing any statistic from the matched sample to a
reference built from that same sample returns the same verdict however tightly the
design matched. Here is why, and a check.

Under within-set re-randomization the covariates are fixed and only treatment is
permuted, so the within-set variance $V$ --- built from the covariate spread and
the fixed treated/control counts, neither of which moves when you reshuffle
treatment --- is the *same* for every permutation. Dividing the observed value and
every permuted value of $d^2$ by that same constant cannot change a rank, so the
percentile of $d^2/V$ is just the percentile of the bare imbalance $d^2$. Replacing
$V$ with any other fixed standardization --- a fixed pool variance, the full-sample
$\mathrm{cov}(x)$ --- divides by a different constant and changes nothing: every
fixed metric returns the percentile of the bare imbalance, which is what collapses.

```{r}
#| label: appendix-metric-cancels
## A fixed pool reference variance (a loose, un-compressed match), held fixed.
pool   <- make_1d(shrink = 1)
Sigma0 <- perm_var(pool$x, pool$mset, pool$z)

## within-set re-randomization distribution of d^2 / (fixed metric). A shared
## inner seed makes both metrics see the SAME permutations, so any difference in
## the resulting p-value would be real, not Monte Carlo noise.
within_ref <- function(x, blk, metric, B = 4000, seed = 1) {
  set.seed(seed); idxs <- split(seq_along(x), blk)
  replicate(B, {
    zp <- numeric(length(x)); for (idx in idxs) zp[sample(idx, 2)] <- 1
    adj_diff(x, blk, zp)^2 / metric
  })
}

appendix <- sapply(c(0.5, 0.05, 0.01), function(sh) {
  d        <- make_1d(shrink = sh)
  Vreal    <- perm_var(d$x, d$mset, d$z)
  obs_real <- adj_diff(d$x, d$mset, d$z)^2 / Vreal     # observed, released metric
  obs_pool <- adj_diff(d$x, d$mset, d$z)^2 / Sigma0    # observed, fixed metric
  c(shrink      = sh,
    p_realizedV = mean(within_ref(d$x, d$mset, Vreal)  >= obs_real),
    p_fixedPool = mean(within_ref(d$x, d$mset, Sigma0) >= obs_pool),
    magnitude_M = obs_pool)
})
round(t(appendix), 4)
```

The two p-value columns fall together as the match tightens and agree to the last
digit: the metric has cancelled. What the fixed metric *does* change is the size of
the statistic, not its percentile --- the `magnitude_M` column stays put while the
imbalance is held fixed and the noise shrinks, which is why a fixed pool variance
makes a useful precision-invariant *magnitude* to report beside the test, but not a
different test. This is the magnitude the recommendation section reports.

The multivariate case is not an exact cancellation --- a matrix metric can rank
permutations differently --- but the collapse survives it: as the match tightens
the imbalance concentrates in the residual-gap direction and the observed
assignment maximizes that component, so the observed sits at the extreme of the
within-set reference under any metric with weight on that direction.
`vignettes/rescaling-prototype.R` confirms this for the omnibus and extends it to
two proposed rescues --- empirical-Bayes (Stein) pooling of each set's scatter
toward the pool, and a covariance-shrinkage dial. Each keeps the *statistic*
bounded as the match tightens, but leaves the within-set *p-value* collapsing. The
only reference that does not collapse is a coarser one --- the resolution profile of
the optional section above. The competing-statistics table shows the same collapse
in the standard F-test and logistic-regression LR balance tests, the quadratic
form's relatives, and in the Westfall-Young max-$T$ permutation omnibus, which is
not a quadratic form at all --- the sharpest form of the claim here, that the verdict
is a property of the reference, not of the shape of the statistic.

---

## Appendix B. The R-squared relabeling, and why distances from its baselines add no test

An R-squared from regressing the (within-stratum-centered) treatment indicator on
the (within-stratum-centered) covariates is the $d^2$ test on an interpretable
$[0,1]$ scale, not a new test. The exact relation is $R^2 = T^2 / (T^2 + \text{df})$
for a Hotelling/Mahalanobis quadratic form $T^2$ in the same $d$, so $R^2$, the F
statistic, and $d^2$ are monotone transforms of one another. The monotone
equivalence is exact for $K = 1$ or for designs with a constant per-block weight
(pure pair matching, equal-size and equal-ratio designs), and equal to simulation
precision otherwise. Reading $R^2$ against its within-block re-randomization
distribution therefore gives the same percentile as $d^2$, and collapses the same
way.

The interpretable content is the two baselines. The ideal floor is $R^2 = 0$
(covariates constant within strata; treatment can explain nothing). The chance floor
is $K/(n - S)$ with $S$ strata --- approximately the mean of $R^2$ under the
within-block null, since $K$ covariates explain about $K$ of the $n - S$
within-stratum residual degrees of freedom by chance. The natural "how far above
well-balanced" is the distance from the chance floor, normalized, and that is
*exactly* adjusted $R^2$: $\text{adjusted } R^2 = (R^2 - K/(n-S)) / (1 - K/(n-S))$,
an identity.

A randomization distribution *of that distance* adds no test. The baselines are
constants --- they do not depend on the assignment --- and subtracting a constant
preserves ranks, so the within-block percentile of $(R^2 - \text{baseline})$ equals
the percentile of $R^2$, which equals the $d^2$ percentile. The distance is a useful
effect size, not a second, independent check. The distance becomes a live,
non-collapsing object only by changing the metric (a fixed pool metric, giving the
absolute magnitude of the recommendation) or by changing the question (equivalence
within a tolerance). The runnable version of this argument, with the constant-shift
demonstration and the magnitude-versus-calibration contrast, is in
`vignettes/r2-balance-reasoning.qmd`.
