---
title: "Development memo: the sigma_x balance test and its four null backends"
author: "Jake Bowers, Ben Hansen, Mark Fredrickson, Josh Errickson (and Claude)"
date: "Drafted on the `devel-sigma-x-omnibus` branch"
output:
  html_document:
    toc: true
    toc_depth: 3
---

This is a working memo on the development branch `devel-sigma-x-omnibus`. It is
not a polished package vignette and is not built as part of the `RItools`
package distribution (the `vignettes/` directory is in `.Rbuildignore`). Its
purpose is to give the team --- and any readers we share the branch with ---
a concrete sense of when the new omnibus statistic and its four
null-distribution backends agree, when they disagree, and what each
disagreement actually means.

# 1. Background

`balanceTest()` in `RItools` reports a multivariate omnibus test for the
hypothesis of no covariate imbalance between treatment and control. The
released version uses the Hansen-Bowers (2008) statistic

$$T_{HB} \;=\; d'\,\mathrm{Cov}(d)^{-1}\,d,$$

where $d$ is the (stratification- and weight-aware) vector of treatment-vs-
control mean differences across the $p$ covariates and $\mathrm{Cov}(d)$ is
the covariance of $d$ under the strata-conditional randomization
distribution. Under that randomization null and an asymptotic-normal
approximation for $d$, $T_{HB}$ is asymptotically chi-squared with degrees of
freedom equal to the rank of $\mathrm{Cov}(d)$.

This branch adds an alternative statistic

$$T_{\Sigma_x} \;=\; d'\,\Sigma_x^{-1}\,d,$$

where $\Sigma_x$ is *not* the permutation covariance but rather a covariance
of the *covariates themselves*. By default $\Sigma_x$ is the within-stratum-
pooled sample covariance $(N-K)^{-1}\sum_s\sum_{i\in s}(x_i - \bar x_s)(x_i -
\bar x_s)'$; the user may also supply any positive semidefinite matrix
(e.g. a covariance computed from a larger reference population not affected
by the matching).

Under the same randomization null, the exact distribution of $T_{\Sigma_x}$
is no longer chi-squared. Asymptotically it is

$$T_{\Sigma_x} \;\sim\; \sum_{k=1}^{r} \lambda_k \,\chi^2_{1,k},
\qquad \lambda_k = \text{eigenvalues of }\Sigma_x^{-1/2}\,V_d\,\Sigma_x^{-1/2},$$

a weighted mixture of independent $\chi^2_1$ random variables. At finite $n$
the exact distribution differs from this mixture --- and the gap can matter
in practice.

The branch implements four ways of computing a p-value for $T_{\Sigma_x}$,
selected via the `null` argument to `balanceTest()`:

| Backend | Assumption | Cost | Honest at finite $n$? |
|---|---|---|---|
| `satterthwaite_asymptotic` | Gaussian $d$, moments via $E[T]=\sum\lambda_k$, $\mathrm{Var}[T]=2\sum\lambda_k^2$ | $O(p^3)$ | No (conservative when $T$ is in upper tail) |
| `satterthwaite_finite` | Chi-square *shape*, moments via the *exact* within-stratum randomization distribution | $O(p^2 n)$ | Closer, but still wrong when shape itself fails |
| `imhof` | Gaussian $d$, exact CDF inversion via `CompQuadForm` | $O(p)$ per evaluation | No (same Gaussian-$d$ assumption as `satterthwaite_asymptotic`, just no shape approximation) |
| `simulate` | None (Monte Carlo over the within-stratum randomization distribution) | $O(B \cdot np)$ | **Yes**, given enough $B$ |

The first three are the cheap analytic options. The fourth is slow but
asymptotically (as $B \to \infty$) gives the exact randomization-null
p-value.

# 2. When do the backends agree?

```{r setup, message=FALSE, warning=FALSE}
devtools::load_all("..", quiet = TRUE)
data(nuclearplants)
```

## A trivial cancellation: unstratified, default $\Sigma_x$

In the unstratified case with the default $\Sigma_x = (n-1)^{-1}S_x$, the
permutation covariance is $V_d = (n_1 n_0/(n(n-1))) S_x$, so

$$\Sigma_x^{-1}\,V_d \;=\; \frac{n_1 n_0}{n}\,I_p$$

and *every* eigenvalue equals $n_1 n_0 / n$. The mixture $\sum_k \lambda_k
\chi^2_1$ collapses to $(n_1 n_0/n)\,\chi^2_p$, and $T_{\Sigma_x} =
(n_1n_0/n)\cdot T_{HB}$. The *p*-value of $T_{\Sigma_x}$ under the asymptotic
backends is exactly the chi-square p-value of $T_{HB}$.

```{r unstrat-collapse}
bt <- balanceTest(pr ~ date + t1 + t2 + cap + ne + ct + bw + cum.n,
                  data = nuclearplants, sigma_x_test = TRUE,
                  null = "satterthwaite_asymptotic")
bt$overall["--", c("chisquare", "p.value", "sigma_x", "sigma_x.p.value")]
## sigma_x.p.value should equal p.value to machine precision
```

The same cancellation does **not** happen with `satterthwaite_finite`,
because the actual randomization variance $\mathrm{Var}[T]$ is smaller than
the Gaussian-derived $2\sum\lambda_k^2$ at finite $n$. So even in this
trivial-looking case the finite-sample backend reports a different (and
slightly smaller) p-value.

## With stratification: the four backends start disagreeing

As soon as we stratify, $\Sigma_x \not\propto V_d$ (because the within-
stratum-pooled covariance weights strata equally while the permutation
covariance weights them by harmonic counts), and the asymptotic and finite-
sample backends part company.

```{r strat-disagree}
bt <- balanceTest(pr ~ date + t1 + t2 + cap + ne + ct + bw + cum.n + strata(pt),
                  data = nuclearplants, sigma_x_test = TRUE)
## We have to call balanceTest() three more times to get the other backends
fmla <- pr ~ date + t1 + t2 + cap + ne + ct + bw + cum.n + strata(pt)
ps <- sapply(c("satterthwaite_finite", "satterthwaite_asymptotic",
               "imhof", "simulate"),
             function(nm) {
               set.seed(20260411)
               balanceTest(fmla, data = nuclearplants, sigma_x_test = TRUE,
                           null = nm, n_simulate = 5000)$overall["pt",
                                                                 "sigma_x.p.value"]
             })
ps
```

# 3. Engineered example A: small $n$ where the chi-square *shape* breaks

This is the cleanest case for showing why even the finite-sample
Satterthwaite backend can disagree with the exact randomization null. We use
a small subset of the nuclearplants data --- only 14 plants, only 3 of them
treated --- so that the full randomization distribution of $T$ has only
$\binom{14}{3} = 364$ points and we can enumerate it exactly.

```{r small-n-setup}
small <- nuclearplants[1:14, ]
fmla  <- pr ~ date + t1 + t2 + cap

## The four analytic backends
ps <- sapply(c("satterthwaite_finite", "satterthwaite_asymptotic", "imhof"),
             function(nm)
               balanceTest(fmla, data = small, sigma_x_test = TRUE,
                           null = nm)$overall["--", "sigma_x.p.value"])

## Simulate at increasing B with fresh seeds
sim_ps <- sapply(c(2000, 5000, 20000, 100000), function(B) {
  set.seed(20260411)
  balanceTest(fmla, data = small, sigma_x_test = TRUE,
              null = "simulate", n_simulate = B)$overall["--", "sigma_x.p.value"]
})
names(sim_ps) <- paste0("simulate B=", c(2000, 5000, 20000, 100000))

## And exact enumeration --- the gold standard
X      <- as.matrix(small[, c("date", "t1", "t2", "cap")])
strata <- factor(rep("--", nrow(X)))
Sx     <- default_sigma_x(X, strata)
n      <- 14;  n1 <- sum(small$pr)
T_obs  <- {
  pi <- mean(small$pr)
  d  <- drop((small$pr - pi) %*% X)
  drop(d %*% solve(Sx) %*% d)
}
all_z   <- combn(n, n1)
T_null  <- apply(all_z, 2, function(loc) {
  z <- numeric(n);  z[loc] <- 1
  d <- drop((z - mean(z)) %*% X)
  drop(d %*% solve(Sx) %*% d)
})
exact_p <- mean(T_null >= T_obs)

c(ps, sim_ps, exact = exact_p)
```

The numerical pattern (verified on the canonical run): all three analytic
backends report ~0.338, simulate converges to ~0.383, and the exact
enumeration returns 0.382. **The analytic backends are off by 0.044 in
absolute terms** --- a real bias, not Monte Carlo noise.

Why? Look at the moments:

```{r small-n-moments}
EM_exact   <- mean(T_null)
VarM_exact <- mean((T_null - EM_exact)^2)
mom        <- sigma_x_T_moments(X, strata,
                                n1_per_stratum = n1, sigma_x = Sx)
data.frame(quantity = c("E[T]", "Var[T]"),
           exact_enumeration = c(EM_exact, VarM_exact),
           sigma_x_T_moments = c(mom$EM, mom$VarM))
```

The finite-sample backend has both moments **exactly right** (down to
machine precision). What it gets wrong is the *shape* --- it forces the
distribution into a $a\cdot\chi^2_v$ template, but the exact distribution at
$n=14$ is more skewed and longer-tailed than any chi-square can capture. At
this sample size the chi-square *shape assumption* is the binding constraint,
not the moment match.

```{r small-n-skew}
## Empirical higher moments of the exact null
sk <- function(x) mean(((x - mean(x))/sd(x))^3)
ku <- function(x) mean(((x - mean(x))/sd(x))^4) - 3
v  <- 2 * EM_exact^2 / VarM_exact      # finite-sample df
data.frame(
  quantity                    = c("skewness", "excess kurtosis"),
  exact_null                  = c(sk(T_null),     ku(T_null)),
  matched_chi_square_v_6.7    = c(sqrt(8 / v),    12 / v)
)
```

The chi-square at the matched $df = 6.7$ predicts skewness $\approx 1.09$
and excess kurtosis $\approx 1.78$. The exact null distribution has very
different higher moments. Two-moment matching is not enough at this $n$.

**Lesson:** at small $n$, even the finite-sample Satterthwaite backend is
biased. Use `simulate` (or, when feasible, exact enumeration) when the
effective sample size is small. The cost of `simulate` is linear in $B$ and
the per-draw cost is small, so $B = 50,000$ on a 14-unit example is essentially
instant.

# 4. Engineered example B: tight matching with persistent imbalance (Rosenbaum's criticism)

This example targets Paul Rosenbaum's concern (relayed by Ben Hansen):

> "If your pairing vastly reduced variation in $x$, but at the same time
> left a meaningful treatment-control imbalance on $x$, we would tend to
> reject it for that reason. His feeling was that this wouldn't be a good
> reason to reject that setup."

The key dynamic: under tight matching, the within-stratum residual scale
$\varepsilon$ shrinks. The numerator of $T_{HB}$ (i.e. $d$) does *not*
necessarily shrink with $\varepsilon$, because the treatment-control
difference can persist. The denominator $\mathrm{Cov}(d)$ scales as
$\varepsilon^2$. So the ratio $T_{HB} = d'\mathrm{Cov}(d)^{-1}d$ scales as
$\varepsilon^{-2}$ if $d$ is held fixed --- it explodes as the matching gets
tighter, regardless of whether the absolute imbalance is meaningful in
covariate-natural units.

Critically, the **default** $\Sigma_x$ in the new test inherits the same
problem: the within-stratum-pooled covariance also scales as $\varepsilon^2$.
$T_{\Sigma_x}$ with the default $\Sigma_x$ also explodes. The choice of
*null backend* does not save us --- the issue is in the *statistic*, not in
how we compute its reference distribution.

What addresses the criticism is a **design-independent** $\Sigma_x$: a
matrix of natural-scale covariate variances that is *not* recomputed from
the matched sample. The cleanest example is a covariance from a reference
population. The `sigma_x` argument lets the user supply such a matrix.

```{r rosenbaum-fixture}
## Construct a 4-stratum design where matching has tightened within-stratum
## variance dramatically, but a small persistent treatment-control imbalance
## remains. We work with two covariates for clarity.
set.seed(20260413)
n_per_s <- 4L
K       <- 4L
n       <- n_per_s * K
strata  <- factor(rep(seq_len(K), each = n_per_s))
shrink  <- 0.05                                # within-stratum scale
gap     <- 0.5                                 # T-vs-C mean diff in raw units

## Stratum centers (these are the "true" covariate values, untouched by matching)
centers <- matrix(rnorm(K * 2, sd = 3), K, 2)

## Within each stratum: 2 treated, 2 control, with a persistent gap.
## Within-stratum noise scaled by `shrink`.
X <- matrix(NA_real_, n, 2)
z <- integer(n)
for (s in seq_len(K)) {
  idx <- which(as.integer(strata) == s)
  X[idx, ] <- matrix(centers[s, ], 4, 2, byrow = TRUE) +
              shrink * matrix(rnorm(8), 4, 2)
  z[idx[1:2]] <- 1
  ## Add the gap to treated rows in dim 1 only
  X[idx[1:2], 1] <- X[idx[1:2], 1] + gap
}
colnames(X) <- c("x1", "x2")
df <- data.frame(z = z, X, strata = strata)

## Population covariance: what x1 and x2 look like in the full population,
## not affected by the matching
pop_cov <- diag(c(9, 9))   # var = 9 each (matching the centers' spread)

## Run the four backends with the DEFAULT sigma_x
default_ps <- sapply(c("satterthwaite_finite", "satterthwaite_asymptotic",
                       "imhof", "simulate"),
                     function(nm) {
                       set.seed(20260413)
                       balanceTest(z ~ x1 + x2 + strata(strata), data = df,
                                   sigma_x_test = TRUE, null = nm,
                                   n_simulate = 10000)$overall["strata",
                                                                "sigma_x.p.value"]
                     })

## And the same with the EXTERNAL population sigma_x
external_ps <- sapply(c("satterthwaite_finite", "satterthwaite_asymptotic",
                        "imhof", "simulate"),
                      function(nm) {
                        set.seed(20260413)
                        balanceTest(z ~ x1 + x2 + strata(strata), data = df,
                                    sigma_x_test = TRUE, null = nm,
                                    n_simulate = 10000,
                                    sigma_x = pop_cov)$overall["strata",
                                                                "sigma_x.p.value"]
                      })

bt_hb <- balanceTest(z ~ x1 + x2 + strata(strata), data = df)
data.frame(
  backend                = c("HB08 chi-square", names(default_ps)),
  default_sigma_x_pval   = c(bt_hb$overall["strata", "p.value"], default_ps),
  external_sigma_x_pval  = c(NA, external_ps)
)
```

The pattern to look for in the table above:

* The HB08 chi-square test and the new test with the *default* $\Sigma_x$
  should both give a small p-value --- both are computed with covariance
  matrices that have been shrunk by the matching, so a modest absolute
  imbalance produces a large standardized statistic.
* The new test with the *external* $\Sigma_x$ (the population covariance,
  which is independent of how tight the matching is) should give a larger
  p-value, because the absolute size of $d$ is small relative to the natural
  scale of the covariates.

If you make `shrink` smaller and smaller (tighter matching) while keeping
`gap` fixed, the gap between the two columns above gets larger and larger.
This is a clean numerical illustration of Rosenbaum's criticism --- and of
the fact that the *statistic*, not the *null backend*, is what determines
whether the criticism applies.

# 5. Engineered example C: clusters where simulate is the only honest option

`ym_long` is a clustered design (`cluster(practice)`); after cluster
aggregation the effective $n$ at the cluster level is much smaller than
the individual-level row count would suggest. This is the regime where the
asymptotic backends are conservative or anti-conservative depending on
where the test stat falls in the chi-square body.

```{r ym-long-cluster}
data(ym_long)
fmla <- trt ~ assessed + hypo + lipid + aspirin +
              strata(assess_strata) + cluster(practice)
ps <- sapply(c("satterthwaite_finite", "satterthwaite_asymptotic",
               "imhof", "simulate"),
             function(nm) {
               set.seed(20260411)
               balanceTest(fmla, data = ym_long, sigma_x_test = TRUE,
                           null = nm,
                           n_simulate = 5000)$overall["assess_strata",
                                                      "sigma_x.p.value"]
             })
ps
```

# 6. Recommendations from this memo (subject to team review)

1. **Default null backend.** `satterthwaite_finite` is the right cheap
   default. It gets the moments right even when the asymptotic-Gaussian
   identities fail, and it collapses to the asymptotic backends in the
   large-$n$ limit. Document the limitation that even it can be biased at
   very small $n$ because of the chi-square *shape* assumption.

2. **When to switch to `simulate`.** Whenever the effective sample size
   (post-aggregation, post-stratification) is small enough that the
   randomization distribution of $T$ has visibly non-Gaussian higher
   moments. As a rough rule of thumb: prefer `simulate` when the total
   number of distinct randomizations is small enough to enumerate, *or*
   when the cluster-level effective $n$ is below ~50. Cost of `simulate`
   at $B = 5000$ is on the order of seconds for typical designs.

3. **Default $\Sigma_x$.** This memo does **not** resolve whether the
   default should remain the within-stratum-pooled sample covariance.
   Rosenbaum's criticism (Section 4) suggests it should not. The two
   alternative defaults to consider:
   * The *full-sample* (pre-stratification) sample covariance of the
     covariates, which is design-independent in the sense that it does not
     shrink with tighter matching.
   * Force the user to supply $\Sigma_x$ explicitly --- no default. This
     would surface the choice as a deliberate methodological decision.

   We are dispatching two independent mathematical-statistics analyses to
   sharpen the recommendation.

4. **Vignette / doc warnings.** Whatever default we pick, the documentation
   for `balanceTest()` should explicitly warn that the test stat is sensitive
   to the choice of $\Sigma_x$ and that, in particular, the default pooled
   covariance can give misleading p-values when matching is tight.

5. **Tests we should add (after the team has decided):** an explicit
   regression test reproducing the small-$n$ chi-square-shape failure on a
   tiny enumerable example, anchored to the gold-standard exact-enumeration
   p-value. We have helpers for this in
   `tests/testthat/helper-sigma_x_test.R`.

# 7. Open questions for the team

1. Should the default `null` argument stay at `satterthwaite_finite`, or
   should we promote `simulate` to the default for safety at the cost of
   speed?
2. Should the default `sigma_x` be changed (see Section 6 item 3)?
3. Should we expose a "compare to complete randomization" backend in the
   spirit of Rosenbaum's 2025 chapter --- i.e., a simulate variant that
   draws z's from complete randomization on $N$ units rather than from
   within-stratum SRSWOR? This is methodologically distinct from what
   `null = "simulate"` currently does.
4. Should we expose an `exact` backend that enumerates the randomization
   distribution when the number of randomizations is small enough to make
   it feasible (say, < 10,000)?
5. Anything else the team wants to test on the way to deciding on a final
   API.
