---
title: "What an R^2 balance summary buys: baselines, distances, and the calibration trilemma"
author: "Jake Bowers (with Claude)"
date: "2026-06-16"
format:
  html:
    toc: true
    embed-resources: true
    code-fold: show
    df-print: kable
---

```{=html}
<!--
This document reasons about whether an R^2-style balance summary, and "distances"
from its natural baselines, give anything beyond the Hansen-Bowers d^2 omnibus.
Every numeric claim below is produced by the code chunk above it; the helper code
was checked with Rscript (2026-06-16). Branch: devel-sigma-x-omnibus. Companions:
r2-and-unchosen-designs-memo.md, resolution-profile-memo.md,
resolution-profile-threebenchmark.R, sigma-x-team-decision-memo.md.
-->
```

## The question on the table

Jake's idea: an R^2 of treatment on covariates is easier to read than a chi-square,
and it carries its own "well balanced" baselines --- the chance level `K/(n-S)` and
the perfect-stratification ideal `R^2 = 0`. Could the *distance* from one or both of
those baselines be the balance measure, perhaps with its own randomization
distribution?

The short answers, established below with code: (1) `R^2` is the d^2 test on a [0,1]
scale, so its randomization percentile is d^2's. (2) The distance from the chance
floor, normalized, is *exactly* adjusted `R^2` --- a clean effect size. (3) A
randomization distribution *of that distance* gives nothing new, because the
baselines are constants and subtracting a constant cannot change a rank. (4) The
distance becomes a live, non-collapsing object only by changing the metric (a fixed
external scale: absolute magnitude) or by changing the question (equivalence within
a tolerance). (5) All of this is one face of a trilemma that also governs the l*
resolution profile.

## Setup: the arithmetic, matched to the package

These helpers reproduce `randomization_cov_d` / `balanceTestEngine` from scratch so
the d^2 here is the released statistic. `d = sum_i c_i x_i` is the within-block-
adjusted difference (contrast `c_i = z_i - pi_b`); `V_d` is its within-block
permutation covariance; `Sx` is the within-block-centered covariate scatter (the
regression metric behind `R^2`).

```{r setup}
center_contrast <- function(raw, blk) raw - ave(raw, blk, FUN = mean)
adj_diff <- function(X, blk, cc) drop(crossprod(cc, as.matrix(X)))

## within-block-centered covariate scatter (the regression metric for R^2)
Sx_scatter <- function(X, blk) {
  X <- as.matrix(X); S <- matrix(0, ncol(X), ncol(X))
  for (idx in split(seq_len(nrow(X)), blk)) {
    if (length(idx) < 2) next
    Xc <- scale(X[idx, , drop = FALSE], scale = FALSE); S <- S + crossprod(Xc)
  }
  S
}
## d^2 randomization covariance V_d = sum_b ( sum_i c_i^2 / (n_b-1) ) Sx_b
Vperm <- function(X, blk, cc) {
  X <- as.matrix(X); V <- matrix(0, ncol(X), ncol(X))
  for (idx in split(seq_len(nrow(X)), blk)) {
    n <- length(idx); if (n < 2) next
    sc <- sum(cc[idx]^2); if (sc <= 0) next
    Xc <- scale(X[idx, , drop = FALSE], scale = FALSE); V <- V + (sc/(n-1)) * crossprod(Xc)
  }
  V
}
## q = d' M^+ d with HB08 eigen rank truncation; returns q and rank
quad_pinv <- function(d, M, tol = sqrt(.Machine$double.eps)) {
  M <- (M + t(M))/2; e <- eigen(M, symmetric = TRUE)
  pos <- e$values > tol * max(e$values, 1)
  if (!any(pos)) return(c(q = 0, rank = 0))
  proj <- crossprod(e$vectors[, pos, drop = FALSE], d)
  c(q = unname(sum(proj^2 / e$values[pos])), rank = sum(pos))
}
d2_val <- function(X, blk, z) { cc <- center_contrast(z, blk)
  unname(quad_pinv(adj_diff(X, blk, cc), Vperm(X, blk, cc))["q"]) }
r2_val <- function(X, blk, z) { cc <- center_contrast(z, blk)
  unname(quad_pinv(adj_diff(X, blk, cc), Sx_scatter(X, blk))["q"]) / sum(cc^2) }

## R^2 with its two baselines and adjusted R^2
r2_stats <- function(X, blk, z) {
  cc <- center_contrast(z, blk); d <- adj_diff(X, blk, cc)
  n <- nrow(as.matrix(X)); S <- nlevels(factor(blk))
  ess <- quad_pinv(d, Sx_scatter(X, blk)); Keff <- unname(ess["rank"])
  R2 <- unname(ess["q"]) / sum(cc^2)
  chance <- Keff / (n - S); adjR2 <- 1 - (1 - R2) * (n - S) / (n - S - Keff)
  c(R2 = R2, chance = chance, adjR2 = adjR2, K = Keff)
}

## 4 groups x 4 matched sets x (2T+2C); persistent gap g, within-set noise = shrink
make_data <- function(shrink, g = 0.2, K = 2, group_centers = c(0,10,20,30),
                      sets_per_group = 4) {
  Xl <- list(); ms <- integer(0); z <- integer(0); sid <- 0L
  for (gc in group_centers) for (s in seq_len(sets_per_group)) {
    sid <- sid + 1L; ctr <- gc + rnorm(K, sd = 1.5)
    Xl[[sid]] <- rbind(ctr + g/2 + shrink*rnorm(K), ctr + g/2 + shrink*rnorm(K),
                       ctr - g/2 + shrink*rnorm(K), ctr - g/2 + shrink*rnorm(K))
    ms <- c(ms, rep(sid, 4)); z <- c(z, c(1, 1, 0, 0))
  }
  X <- do.call(rbind, Xl); colnames(X) <- paste0("x", seq_len(K))
  list(X = X, set = factor(ms), z = z)
}
```

## 1. R^2 is the d^2 test on a [0,1] scale

`R^2` from regressing the within-stratum-centered treatment indicator on the
within-stratum-centered covariates equals `T2 / (T2 + df)` for a Mahalanobis
quadratic form `T2` in the same `d`. At a blocking whose blocks share a common
size and treated:control ratio (the matched-set blocking here, all 2T:2C), `R^2`
and d^2 are a strictly increasing relabeling of each other, so their within-block
randomization percentiles are identical.

```{r monotone}
set.seed(303)
d   <- make_data(shrink = 0.15)
blk <- d$set                                   # matched sets, all 2T:2C (homogeneous)

R2o <- r2_val(d$X, blk, d$z); d2o <- d2_val(d$X, blk, d$z)
B <- 4000
R2p <- d2p <- numeric(B)
for (b in seq_len(B)) {
  zp <- d$z; for (idx in split(seq_along(zp), blk)) zp[idx] <- sample(zp[idx])
  R2p[b] <- r2_val(d$X, blk, zp); d2p[b] <- d2_val(d$X, blk, zp)
}
c(spearman_R2_d2 = cor(R2p, d2p, method = "spearman"),
  perc_R2 = (1 + sum(R2p >= R2o - 1e-12)) / (B + 1),
  perc_d2 = (1 + sum(d2p >= d2o - 1e-12)) / (B + 1))
```

The Spearman correlation is 1 and the two percentiles agree to the digit:
`R^2`-against-its-randomization-distribution *is* the d^2 test, relabeled onto
[0,1]. (For full matching with unequal block sizes the relabeling is monotone only
to simulation precision; see r2-and-unchosen-designs-memo.md sec. 1.2.)

## 2. The distance from the chance floor is adjusted R^2

`R^2` carries two baselines. The ideal floor is `R^2 = 0`: if the covariates were
constant within strata there is nothing for treatment to explain. The chance floor
is `K/(n-S)`, which is approximately the *mean* of `R^2`'s own randomization
distribution under within-block randomization --- with `K` covariates and `n-S`
within-stratum residual degrees of freedom, chance fitting explains about `K/(n-S)`
even when as-if-randomization holds. The natural "how far above well-balanced is my
design" is the distance from that chance floor, normalized by the room above it,
and that quantity is exactly adjusted `R^2`:

$$
\text{adjusted } R^2 \;=\; \frac{R^2 - K/(n-S)}{1 - K/(n-S)}.
$$

```{r adj-identity}
st <- r2_stats(d$X, blk, d$z)
c(R2 = st["R2"], chance = st["chance"], adjR2 = st["adjR2"],
  normalized_distance = (st["R2"] - st["chance"]) / (1 - st["chance"]),
  identity_gap = unname(st["adjR2"] - (st["R2"] - st["chance"]) / (1 - st["chance"])))
```

The identity gap is 0. So the statistic Jake is reaching for already has a name:
adjusted `R^2` is the design's position above its chance baseline, and it is the
cleanest one-number effect size the R^2 reframing produces. (The chance floor is the
randomization *mean*; the percentile-1/2 crossing used for l* is the *median*, so
adjusted-`R^2 = 0` and percentile `= 1/2` are close but not identical for a skewed
distribution --- which is why the two l* readings in
resolution-profile-threebenchmark.R came out at 10.4 and 11.5.)

## 3. A randomization distribution of the distance adds nothing

Here is the catch. Both baselines --- `K/(n-S)` and `0` --- are *constants*: they
do not depend on the treatment assignment, so they do not move when treatment is
re-randomized within blocks. Form the distance `D(z) = R^2(z) - baseline` and build
its within-block randomization distribution: every value is just `R^2(z)` shifted by
the same constant. Subtracting a constant preserves ranks, so the percentile of the
observed distance equals the percentile of `R^2`, which (Section 1) equals d^2's.

```{r constant-shift}
c0 <- st["chance"]
perc_R2   <- (1 + sum( R2p           >= R2o        - 1e-12)) / (B + 1)
perc_dist <- (1 + sum((R2p - c0)      >= (R2o - c0) - 1e-12)) / (B + 1)
perc_d2   <- (1 + sum( d2p           >= d2o        - 1e-12)) / (B + 1)
c(perc_R2 = unname(perc_R2), perc_distance_from_chance = unname(perc_dist),
  perc_d2 = unname(perc_d2))
```

All three percentiles are the same number. You cannot buy a new test by
re-centering on a fixed point. The baselines are valuable --- they let you read
`R^2` as an effect size *without simulating a reference* (that is the three-benchmark
plot in resolution-profile-threebenchmark.R) --- but they are reference points, not
the seed of a second test.

## 4. Why calibrating the distance brings back the collapse

To turn the distance into a *calibrated* statement ("is a distance this large
surprising?") you need the *spread* of the null distribution, not just its center.
The center you get for free (`K/(n-S)`). The spread of `R^2`'s within-block
randomization distribution shrinks as the match tightens, so standardizing the
distance by it returns d^2 and its collapse. The table below tightens the match
(smaller `shrink`, fixed gap) and shows the within-pair percentile and the absolute
(fixed-metric) magnitude moving in *opposite* directions.

```{r magnitude-vs-calibration}
set.seed(11); Sigma <- cov(make_data(0.6)$X)        # a fixed pre-match-pool metric
within_perc <- function(X, blk, z, B = 2000) {
  d2o <- d2_val(X, blk, z); cnt <- 1L
  for (b in seq_len(B)) { zp <- z
    for (idx in split(seq_along(zp), blk)) zp[idx] <- sample(zp[idx])
    if (d2_val(X, blk, zp) >= d2o - 1e-12) cnt <- cnt + 1L }
  cnt / (B + 1)
}
maha_fixed <- function(X, blk, z, Sig) { cc <- center_contrast(z, blk)
  d <- adj_diff(X, blk, cc); drop(t(d) %*% solve(Sig) %*% d) }

tab <- data.frame(shrink = c(0.6, 0.3, 0.1, 0.03), within_percentile = NA,
                  withinR2 = NA, adjR2 = NA, fixed_magnitude = NA)
for (i in seq_len(nrow(tab))) {
  set.seed(303); dd <- make_data(tab$shrink[i]); b <- dd$set
  tab$within_percentile[i] <- within_perc(dd$X, b, dd$z)
  tab$withinR2[i] <- r2_val(dd$X, b, dd$z)
  tab$adjR2[i] <- unname(r2_stats(dd$X, b, dd$z)["adjR2"])
  tab$fixed_magnitude[i] <- maha_fixed(dd$X, b, dd$z, Sigma)
}
round(tab, 4)
```

As the match tightens, the within-pair percentile collapses toward 0 --- the design
looks *worse* against the standard Jake trusts --- while the fixed-metric magnitude
falls --- the design looks *better* in absolute units. The within-metric `R^2` and
adjusted `R^2` rise with the collapsing percentile, because they are that standard's
effect size, not an absolute one. This is the significance-versus-magnitude crux: a
single number from the within-pair standard cannot say "the imbalance is small,"
because that standard's calibration is about precision, not size.

## 5. Two ways the distance becomes live

The distance-from-baseline idea bites only when you change one of the two things the
constant-shift argument holds fixed.

**Change the metric: distance from 0 in a fixed external scale.** Everything in
Sections 1--4 uses the within-pair-metric `R^2`, whose value is scale-invariant.
Compute `R^2` (or its Mahalanobis numerator) in a *fixed* pre-match-pool metric and
its distance from the ideal floor `0` genuinely shrinks as the match improves --- the
`fixed_magnitude` column above. That is the sigma_x magnitude, and it is the one
quantity that recognizes a trivial imbalance as trivial. If "distance from a
balanced baseline" should mean "how far from perfect balance, in absolute units,"
this is the version that works.

**Change the question: equivalence against the ideal floor.** Instead of asking "is
`R^2` surprisingly far above 0?" (which collapses), ask "is the absolute imbalance
within a tolerance `delta` of 0?" --- is the design close enough to perfect. The
fixed magnitude makes this concrete: pick a margin and read off when the design
clears it.

```{r equivalence-sketch}
delta <- 0.10                                   # tolerance on the fixed-metric magnitude
data.frame(shrink = tab$shrink, fixed_magnitude = round(tab$fixed_magnitude, 3),
           within_tolerance = tab$fixed_magnitude < delta)
```

The conclusion behaves correctly: a tighter match clears the margin, where the
within-pair p-value was collapsing in the opposite direction. The margin `delta` is
where the "how big is too big" judgment lives --- explicit, on a scale you can read,
instead of hidden in a comparison to a design you did not run. (This is a sketch, not
a finished test: a design-respecting equivalence procedure needs a proper null for
the fixed-metric magnitude, which is the sigma_x machinery the team already has on
the branch.)

## 6. The trilemma, and where this leaves the proposal

Sections 3--4 are one face of a single constraint that also governs the l*
resolution profile. Against the within-pair standard Jake trusts, *any* calibration
collapses, because the reference distribution's spread vanishes with the match.
So you cannot have all three of:

1. the within-pair standard (re-randomize treatment within your realized sets),
2. a calibration ("better than X% of randomized versions") that does not collapse,
3. no comparison to designs you did not build.

Pick two. The resolution profile keeps (2) by relaxing (1) --- coarser references
are weaker versions of the same within-similar-units exchangeability assumption,
which is the only honest way to read "as good as a 2:2 experiment." The
distance-from-baseline work above keeps (1) and (3) and gives up (2): it reports a
magnitude (adjusted `R^2`, or the fixed-metric distance), not a non-collapsing
calibration. Equivalence-within-`delta` is the same corner with the magnitude
judgment made explicit.

Where this points the team proposal:

- Make **adjusted `R^2`** (the normalized distance from the chance floor) the
  headline interpretable effect size, beside the d^2 p-value and the per-covariate
  standardized differences. It is descriptive, it is on [0,1], and it needs no
  second simulation.
- Do **not** sell a randomization distribution of the distance as a new test; it is
  the d^2 percentile relabeled (Section 3).
- For "is the absolute imbalance small," use the **fixed-metric magnitude** (the
  sigma_x number) or an **equivalence test** against `delta`. These keep the
  within-pair standard and never invoke a design you did not run.
- Treat the **l\* resolution profile** as an optional sensitivity analysis on the
  *resolution* of the exchangeability assumption --- and only if that relaxation is a
  question you find useful. If it is not, report the within-pair comparison plus the
  magnitude and stop.

The open thread Jake flagged --- coarsening by combining each matched set with its
*nearby* sets on the matching distance scale (a set's location summarized by, say,
the average or median within-set distance) --- is exactly the geometry-respecting
ladder in resolution-profile-threebenchmark.R, and whether that relaxation is worth
making is the next thing to think through, not a settled point.
```
