---
title: "A balance test that rewards careful matching"
subtitle: "On a criticism of HB08 and what it means for the new sigma_x test"
author: "Jake Bowers (with Claude, in conversation with Ben Hansen and Paul Rosenbaum)"
date: today
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
---

# What this memo is about

Paul Rosenbaum has criticized the Hansen-Bowers (2008) balance test on
the following grounds, paraphrased by Ben:

> If your pairing vastly reduced variation in x, but at the same time left
> a meaningful treatment-control imbalance on x, we would tend to reject it
> for that reason. His feeling was that this wouldn't be a good reason to
> reject that setup.

I want to know two things. First: is this criticism right? Second: does the
new test we are building on the `devel-sigma-x-omnibus` branch fix it?

The answers, as best I can tell after running two independent analyses
through AI agents and verifying the math numerically, are: yes, the
criticism is right; and no, the new test as currently coded does not fix
it. There are three things that would fix it, and we should pick one
before this branch becomes a merge candidate. This memo lays out the
problem with a small worked example, shows what the math is doing under
the hood, and proposes a fix.

# Setup

```{r}
#| label: setup
#| message: false
#| warning: false
devtools::load_all("..", quiet = TRUE)
```

# A small worked example

Build a matched study with eight units, in two strata of four. Two
covariates, $x_1$ and $x_2$. In each stratum, three of the units are
controls and they all sit at exactly the same $(x_1, x_2)$ value. The
fourth unit is treated, and it sits a small distance $\delta$ away from
the controls --- in the $x_1$ direction in stratum 1, and in the $x_2$
direction in stratum 2. (Offsetting in different directions in the two
strata keeps $\mathrm{cov}(X)$ full rank, so the math below does not
have to worry about pseudoinverses.)

```{r}
#| label: make-toy
make_toy <- function(delta) {
  X <- rbind(c(10, 0), c(10, 0), c(10, 0), c(10 + delta, 0),
             c(10, 8), c(10, 8), c(10, 8), c(10, 8 + delta))
  data.frame(z      = c(0, 0, 0, 1, 0, 0, 0, 1),
             x1     = X[, 1],
             x2     = X[, 2],
             strata = factor(rep(1:2, each = 4)))
}
make_toy(delta = 0.5)
```

The parameter $\delta$ controls how far the treated unit sits from its
controls within each stratum. It is the knob we will turn throughout
this memo to ask: "if the match gets tighter (smaller $\delta$), does
the balance test notice?" A balance test that rewards careful matching
should give a more encouraging answer (a larger p-value) when $\delta$
is smaller. A balance test that is blind to absolute imbalance will give
the same answer regardless of $\delta$.

When $\delta = 0.5$, the treated unit in each stratum is half a unit
away from its controls along one coordinate. When $\delta = 0.001$,
the treated unit is 0.001 away --- agreement to three decimal places,
in covariates whose range is 0 to 20. A researcher looking at this
design would say "this is well balanced, let's proceed to the outcome
analysis." A balance test should agree.

At any positive $\delta$, $\mathrm{cov}(X)$ is full rank: one large
eigenvalue driven by the between-stratum separation of the cluster
centers (around 18), and one small eigenvalue driven by the within-
stratum offset of the single treated unit in each stratum (which scales
with $\delta^2$).

```{r}
#| label: cov-X-rank
sapply(c(1, 0.5, 0.1, 0.01), function(d) {
  Sx <- cov(as.matrix(make_toy(d)[, c("x1", "x2")]))
  c(evals = sort(eigen(Sx)$values), det = det(Sx))
})
```

# What HB08 says

```{r}
#| label: hb08-vs-delta
deltas <- c(1, 0.1, 0.01, 0.001)
hb_pvals <- sapply(deltas, function(d) {
  bt <- balanceTest(z ~ x1 + x2 + strata(strata), data = make_toy(d))
  c(chisquare = bt$overall["strata", "chisquare"],
    p.value   = bt$overall["strata", "p.value"])
})
colnames(hb_pvals) <- paste0("delta=", deltas)
round(hb_pvals, 4)
```

The test gives the same chi-square statistic --- exactly 6 --- and the
same p-value --- 0.0498 --- regardless of $\delta$. Whether the imbalance
is meaningful ($\delta = 1$) or essentially zero ($\delta = 0.001$),
HB08 returns the same answer, and that answer is "reject at the 5 percent
level."

This is a sharper version of Rosenbaum's complaint than I expected. It
isn't that HB08 *over-rejects* tight matches in some weak statistical
sense. It is that HB08 *literally cannot tell the difference* between a
$\delta$ of 1 and a $\delta$ of 0.001 on this design. The information
about absolute imbalance has been thrown away by the test.

# Why does this happen?

The HB08 statistic is

$$T_{HB} \;=\; d^\top \, \mathrm{Cov}(d)^{-1} \, d,$$

where $d$ is the vector of treated-minus-control mean differences across
the covariates, and $\mathrm{Cov}(d)$ is the covariance of $d$ under the
assumption that the matched design was actually a randomized experiment
within strata. The test asks: is $d$ large compared to the random
fluctuation we would see if treatment were assigned at random within
strata?

The trouble is that *both* parts of that question shrink as the match
tightens. In our toy example, $d$ scales with $\delta$ (because the
treated unit is $\delta$ away from the controls, so the mean difference
is some fraction of $\delta$). And $\mathrm{Cov}(d)$ scales with
$\delta^2$ (because variance is the square of a length-like quantity).
When you divide a quantity that scales as $\delta$ squared by a quantity
that scales as $\delta^2$, you get something that does not depend on
$\delta$ at all. Both numerator and denominator melt away at the same
rate. The result is a constant, and that constant is 6.

A statistician would say "the test is invariant under the perturbation
$\delta \to c\delta$ for any $c > 0$." A plain-English version of the same
thing: the test does not see absolute imbalance; it only sees the
*relative* size of the imbalance compared to the within-stratum spread.
When the within-stratum spread is also small, a small imbalance can look
relatively large. From the test's point of view, our $\delta = 0.001$
design and our $\delta = 1$ design are the same design, scaled by a
positive constant. And HB08 is invariant to positive scaling.

This is what Rosenbaum is objecting to. He is not objecting to the math
of HB08 --- it correctly tests the null hypothesis "the assignments came
from a stratified random experiment." He is objecting to the *use* of
HB08 as a measure of design quality. A good match should be rewarded for
producing absolute balance, not penalized for producing a within-stratum
spread that is small relative to the residual imbalance. HB08 cannot
reward absolute balance because it cannot see absolute balance.

# Does the new test fix this?

The new test on the `devel-sigma-x-omnibus` branch is

$$T_{\Sigma_x} \;=\; d^\top \, \Sigma_x^{-1} \, d,$$

with the same $d$ but a different denominator. Instead of the
randomization covariance $\mathrm{Cov}(d)$, we standardize by some
covariance of the covariates themselves, called $\Sigma_x$. The default
choice in the current branch is the within-stratum-pooled sample
covariance: roughly, the ordinary covariance you would compute from the
covariates after centering each stratum at its own mean.

The hope was that switching from the randomization covariance to a
covariate-based covariance would address the criticism. It does not.

```{r}
#| label: new-test-default
new_default_pvals <- sapply(deltas, function(d) {
  bt <- balanceTest(z ~ x1 + x2 + strata(strata), data = make_toy(d),
                    sigma_x_test = TRUE,
                    null = "satterthwaite_asymptotic")
  c(sigma_x   = bt$overall["strata", "sigma_x"],
    p.value   = bt$overall["strata", "sigma_x.p.value"])
})
colnames(new_default_pvals) <- paste0("delta=", deltas)
round(new_default_pvals, 4)
```

Same story. The new test gives the same statistic (9 instead of 6, but a
constant) and the same p-value (0.0143) for every $\delta$. The reason is
that the within-stratum-pooled sample covariance is built out of the
same within-stratum sums of squares that drive $\mathrm{Cov}(d)$. Both
shrink at the same rate as the match tightens. We swapped one denominator
that melts away with $\delta^2$ for another denominator that melts away
with $\delta^2$.

In fact, in any matched-pair design (every stratum has exactly two units),
the new test with the default $\Sigma_x$ equals the old test, exactly, up
to a known constant. They are literally the same test, multiplied by
$2/K$ where $K$ is the number of pairs. Here is the verification on a
4-pair design:

```{r}
#| label: pair-equivalence
set.seed(20260420)
K <- 4L
pair_data <- data.frame(
  z      = rep(c(1, 0), K),
  x1     = rnorm(2 * K),
  x2     = rnorm(2 * K),
  strata = factor(rep(seq_len(K), each = 2))
)
bt_pair <- balanceTest(z ~ x1 + x2 + strata(strata), data = pair_data,
                       sigma_x_test = TRUE,
                       null = "satterthwaite_asymptotic")
T_HB  <- bt_pair$overall["strata", "chisquare"]
T_new <- bt_pair$overall["strata", "sigma_x"]
c(T_HB = T_HB, T_new = T_new, ratio = T_HB / T_new, expected = 2 / K)
```

The ratio is exactly 0.5, as predicted by $2/K = 2/4$. The new test in
its current form is not a new test in pair-matched designs. It is the
old test wearing a hat.

# What is going on, more deeply: studentization

When I first wrote this memo I thought the fix was to change $\Sigma_x$
to the full-sample covariance $\mathrm{cov}(X)$ (the ordinary covariance
across all units, ignoring strata). The argument was: $\mathrm{cov}(X)$
is dominated by *between*-stratum variation (the differences between
stratum centroids), and tightening the match leaves the stratum centroids
alone, so the test's denominator would no longer shrink with the match.
That argument is half right and half wrong. It is right about the test
statistic; it is wrong about the p-value, and the difference matters.

Let me trace the actual diagnostic. With $\Sigma_x = \mathrm{cov}(X)$:

```{r}
#| label: full-sample-trace
deltas2 <- c(1, 0.5, 0.1, 0.01)
trace_full <- sapply(deltas2, function(d) {
  df <- make_toy(d)
  X  <- as.matrix(df[, c("x1", "x2")])
  Sx <- cov(X)
  bt <- balanceTest(z ~ x1 + x2 + strata(strata), data = df,
                    sigma_x_test = TRUE, sigma_x = Sx,
                    null = "satterthwaite_asymptotic")
  c(sigma_x = bt$overall["strata", "sigma_x"],
    p.value = bt$overall["strata", "sigma_x.p.value"])
})
colnames(trace_full) <- paste0("delta=", deltas2)
round(trace_full, 6)
```

The test statistic *does* shrink with $\delta$ --- it goes from 0.078 at
$\delta = 1$ down to 7.87e-6 at $\delta = 0.01$. So the statistic itself
is now reflecting absolute imbalance. But the p-value is essentially
constant at 0.0143. The fix did not fix the p-value.

This is because the p-value compares the statistic to a reference
distribution whose scale also shrinks with $\delta$. Specifically: the
reference distribution is built from $\mathrm{Cov}(d)$, the variance of
$d$ under stratified randomization. And $\mathrm{Cov}(d)$ is itself a
function of the within-stratum sums of squares of the matched sample.
When the match tightens, those sums of squares shrink, $\mathrm{Cov}(d)$
shrinks, and the eigenvalues that determine the comparison distribution
shrink in lock-step with the statistic. The ratio "statistic over scale"
stays exactly the same, and so does the p-value.

This is what statisticians mean by **studentization**, and it is the
single most important point in this memo. A studentized statistic is one
that has been divided by a sample-based estimate of its own standard
error. The classic example is the t-statistic: divide a mean difference
by an estimated standard error, and the result has a Student-t
distribution that does not depend on the unknown population standard
deviation. That is the *point* of studentization in classical
statistics --- it lets you test without knowing the population scale,
because both numerator and denominator scale with the same unknown
quantity, and the ratio washes out the dependence.

HB08 is a studentized statistic. The "estimated standard error" is
$\mathrm{Cov}(d)$ under stratified randomization, which we know exactly
from the design --- but "knowing" it does not make it independent of the
data. It is still a function of the realized $X$, and it scales with
$X$. When we tighten the match (shrink $X$ within strata), both $d$ and
$\mathrm{Cov}(d)$ shrink at the same rate, and the test statistic
$d' \mathrm{Cov}(d)^{-1} d$ stays put. That is the criticism, in two
words: HB08 is studentized, and a balance test should not be.

The user's question asked exactly the right thing: if we already know
the variance from randomization inference, what could "studentized" even
mean? The answer is that "knowing the variance" still leaves the
variance as a function of the matched data. The variance moves with the
data, the statistic moves with the data, and the ratio is invariant.
Calling the variance "known" is a statement about the mathematical
recipe for computing it (we don't need to estimate a parameter); it is
not a statement about whether the variance is fixed across designs.

# Can complete randomization save us? (No, not on this toy)

My first instinct, and the agents' first instinct, was: if the problem
is that the reference distribution shrinks with the match, switch the
reference distribution to one that doesn't --- complete randomization on
the same units, ignoring the matching. Under complete randomization,
$\mathrm{Cov}(d)$ involves the *total* sum of squares of $X$, not the
within-stratum sum of squares. In this toy, the total sum of squares is
dominated by the between-stratum separation (around 18 in the larger
eigenvalue of $\mathrm{cov}(X)$) and barely moves with $\delta$. So
this ought to fix the pathology.

It doesn't. Not on this toy.

```{r}
#| label: complete-randomization-enumeration
## For each delta: enumerate all C(8, 2) = 28 ways to pick 2 of 8 treated,
## ignoring strata. Compute T = d' cov(X)^{-1} d for each, where d uses the
## within-stratum treated-proportion formula (same formula HB08 uses). Report
## the rank of the observed T inside its own 28-value reference distribution.
enumerate_complete <- function(delta) {
  df     <- make_toy(delta)
  X      <- as.matrix(df[, c("x1", "x2")])
  strata <- df$strata
  Sx     <- cov(X);  Sinv <- solve(Sx)

  z_obs  <- df$z
  pi_obs <- ave(z_obs, strata)
  d_obs  <- drop(crossprod(z_obs - pi_obs, X))
  T_obs  <- drop(d_obs %*% Sinv %*% d_obs)

  combos <- combn(8, 2)
  Tvals  <- apply(combos, 2, function(loc) {
    z <- numeric(8); z[loc] <- 1
    pi_i <- ave(z, strata)
    d <- drop(crossprod(z - pi_i, X))
    drop(d %*% Sinv %*% d)
  })
  c(T_obs = T_obs,
    rank  = sum(Tvals >= T_obs),
    p     = mean(Tvals >= T_obs))
}
sapply(c(1, 0.5, 0.1, 0.01, 0.001), enumerate_complete)
```

Every delta gives rank 1 out of 28 --- the observed assignment is the
*most* extreme under complete randomization, uniquely. The p-value is
$1/28 = 0.036$, constant. So the Rosenbaum chapter 6 prescription, at
least for a test stat of Mahalanobis-in-$\mathrm{cov}(X)$ form, does not
rescue us from the pathology on this toy.

Why not? The same reason as before, just in slightly different clothing.
When I enumerate the 28 complete-randomization assignments and compute
$d$ for each, every one of the 28 $d$ values is a linear function of the
matched data, and in particular every one is a linear function of the
within-stratum differences that scale with $\delta$. The observed $d$
is one of those 28 and it scales with $\delta$. The "null" distribution
is the empirical distribution of the other 27. All 28 quantities share
the same $\delta$ dependence. The rank is preserved.

More generally: **any quantity computed from the matched sample alone
is a function of $\delta$, and any comparison between such quantities
is invariant to rescaling of $\delta$.** This is the real, unfixable
form of the criticism. It is not about the within-stratum reference
distribution, or the choice of $\Sigma_x$, or the null backend. It is
structural: no linear or quadratic statistic in $X$, compared to any
distribution derived from the same $X$, can distinguish "the matched
sample with $\delta = 1$" from "the matched sample with $\delta =
0.001$." The two designs look identical under rescaling, and every
rescaling-invariant test gives them the same answer.

# Why no test on the matched sample alone can see absolute imbalance

This section states the structural finding precisely, because the team
will need to decide what to build on it. I will start with the idea in
plain language, then state it formally, then show it numerically.

## The idea in plain language

Here is the core observation. In our toy, the only way the data changes
as $\delta$ varies is that everything within each stratum gets stretched
or compressed by the same factor. The three controls and the one treated
unit in each stratum all move closer to (or further from) their stratum
center, proportionally. The relative *arrangement* of the four units ---
which one is the outlier, which three are tied, which one is farthest
from the rest --- stays exactly the same. Only the absolute distances
change.

A balance test computed on this matched sample has to compare the
observed assignment to a set of hypothetical assignments on the same
units. Under any of the hypothetical assignments, the test statistic is
also computed from the same $X$. So every hypothetical value rescales
by the same factor as the observed value. The observed test stat might
be the largest, or the second-largest, or the smallest among the
hypothetical values --- but its *position in the list* does not change
when everything rescales together.

This is like ranking runners in a race, and then replaying the race on
a track that is twice as long. Every runner's time doubles, but the
winner is still the winner. The ranking is preserved because all the
times scaled by the same factor.

A p-value is just the position of the observed statistic in the list
of hypothetical values. If the position does not change, the p-value
does not change.

## Formal statement

The paragraph above can be made precise as follows. Two independent
analyses (dispatched as AI agents, verified numerically) converged on
the same statement.

**Setup.** We have a matched sample with covariates $X$, treatment
assignment $z$, and strata $s$. We compute a test statistic $T(X, z)$
and a p-value by comparing $T(X, z_{\text{obs}})$ to the set of values
$\{T(X, z') : z' \in \mathcal{Z}\}$, where $\mathcal{Z}$ is some set of
hypothetical assignments (within-stratum permutations, complete
randomizations, etc.).

**Condition.** Suppose that when we rescale the within-stratum deviations
by a positive constant --- replacing each unit's value $x_i$ with
$\bar x_s + \delta (x_i - \bar x_s)$ for some $\delta > 0$ --- the test
statistic at *every* assignment $z$ gets multiplied by the same factor.
Formally: $T(X_\delta, z) = g(\delta) \cdot T(X_1, z)$ for some
function $g$ that is positive and does not depend on which $z$ we
are evaluating.

In other words: the scaling of $T$ is the same whether the treated
units happen to be the "outliers" or the "inliers." It does not matter
*which* units are treated; the rescaling hits every hypothetical
assignment equally.

**Conclusion.** Under this condition, the p-value does not change with
$\delta$. The proof is one line: multiplying every number in a list by
the same positive constant does not change anyone's position in the list.

**Example.** For HB08's $T = d' \mathrm{Cov}(d)^{-1} d$, the rescaling
factor is $g(\delta) = 1$ (constant). The test statistic is literally
the same number at every $\delta$. For the squared Euclidean distance
$T = \|d\|^2$, the rescaling factor is $g(\delta) = \delta^2$. The
test statistic changes, but every hypothetical value changes by
$\delta^2$ too, so the ranking is preserved.

**What about rank-based statistics?** They have an even stronger
version of the same property: the ranking of units within each stratum
is unchanged by *any* stretching or compression, not just proportional
rescaling. Wilcoxon rank sums, Stephenson rank scores, and any test
built from ranks simply do not see $\delta$ at all.

## What does break the invariance?

The condition above fails when the test statistic imports a fixed
yardstick from outside the matched sample --- something whose value does
not change when $\delta$ changes. Three sources of such a yardstick:

1. **A unit of measurement, fixed at data-collection time.** If $x_1$ is
   measured in years, then $|d| = 0.001$ is "one day" and $|d| = 1$ is
   "one year." The scale lives in the variable's unit. A balance test
   that reports $|d|$ in the original units, together with a rule of
   thumb (e.g., "more than 0.1 SD of the population is a red flag"), is
   sensitive to $\delta$ because "1 year" does not rescale.

2. **An external reference population.** A registry, a census, a
   pre-matching candidate pool. If the user can compute $\Sigma_x$ from
   such a population, the resulting matrix is a fixed quantity that does
   not shrink with the match.

3. **Combining the external scale with a p-value.** If the user supplies
   the external $\Sigma_x$ and we compute $T = d' \Sigma_x^{-1} d$ and
   compare it to a fixed $\chi^2_p$ reference (instead of comparing to a
   permutation distribution on the matched sample), we get a p-value
   that is sensitive to $\delta$.

   There is a reasonable debate about option 3. One view: comparing to
   $\chi^2_p$ with a user-supplied covariance is just effect-size
   reporting dressed up as a hypothesis test, and the user should report
   the effect size directly and skip the p-value. Another view: the
   $\chi^2_p$ comparison is the natural multivariate summary of the
   effect sizes, and providing a combined p-value alongside the per-
   covariate effect sizes is genuinely useful. Both views have merit;
   the package should support both and let the user choose.

All three move the "scale" outside the matched sample. **There is no
option that uses only the matched sample.** The scale invariance is not
a bug in HB08; it is the inevitable consequence of building a test
entirely out of data that can be rescaled without changing any of the
combinatorial relationships between treated and control units.

## What does NOT fix it

For the record, here is the list of things I tried, all of which
preserve the scale invariance:

**Switching among the existing four null backends.** The four current
options (`satterthwaite_asymptotic`, `satterthwaite_finite`, `imhof`,
`simulate`) all compute the p-value of a given statistic against the
within-stratum randomization distribution. Switching backends does not
change the scale-invariance property.

**Changing $\Sigma_x$ to the full-sample covariance $\mathrm{cov}(X)$.**
The test statistic does move slightly with $\delta$ in this case
(because $\mathrm{cov}(X)$ contains a small within-stratum-variation
component that also depends on $\delta$), but the movement is tiny and
the p-value is essentially constant. Verified:

```{r}
#| label: full-sample-doesnt-fix
sapply(c(1, 0.5, 0.1, 0.01, 0.001), function(d) {
  df <- make_toy(d)
  Sx_full <- cov(as.matrix(df[, c("x1", "x2")]))
  bt <- balanceTest(z ~ x1 + x2 + strata(strata), data = df,
                    sigma_x_test = TRUE, sigma_x = Sx_full,
                    null = "satterthwaite_asymptotic")
  c(T_new = bt$overall["strata", "sigma_x"],
    p     = bt$overall["strata", "sigma_x.p.value"])
})
```

**Letting the user supply an external $\Sigma_x$, but still using the
within-stratum randomization null.** Even if $\Sigma_x$ is fixed from
outside, the eigenvalues $\lambda_k$ in the reference distribution are
still computed from $V_d$, which is built from the matched sample.
$V_d$ shrinks with $\delta$, so the reference distribution shrinks, so
the p-value is invariant. This path would work only if the reference
distribution were also decoupled from the matched sample --- which
brings us to option 3 in the previous section (a fixed external null,
such as $\chi^2_p$) rather than the randomization null.

**Rosenbaum's chapter 6 complete-randomization reference.** As the
enumeration above shows, this does not fix the toy either. The
complete-randomization reference *does* help in examples where the
between-stratum variation dominates and the observed $d$ is not in the
most-extreme rank, but it does not fix the pathological family of
designs that motivated this memo.

In short: **no test built entirely out of the matched sample can reward
absolute balance.** The resolution has to come from somewhere else.

## A sharper version of the criticism: even Rosenbaum's benchmark is caliper-dependent

Before leaving the structural impossibility result, I want to sharpen
one thing the memo has been too generous about. I have been describing
Rosenbaum's complete-randomization benchmark as "a reference that
does not depend on the matched data." That framing is not quite right.
Rosenbaum's benchmark does not depend on the *assignment* of treatment
within the matched sample, but it *does* depend on which units are in
the matched sample --- and the set of matched units is itself a
function of design decisions like the choice of caliper, the choice of
distance metric, and the criteria for dropping unmatchable units.
Tighten the caliper, drop some outliers, and $\mathrm{cov}(X)$ computed
on the surviving units shifts. The reference distribution shifts too.
Rosenbaum's benchmark inherits the scale-invariance criticism one level
up: it is invariant to the assignment step, not to the caliper step.

Here is a small demonstration. Take 30 units (15 treated, 15 control),
apply a one-dimensional matching with a caliper that drops treated-
control pairs whose distance exceeds a threshold, and compute
Rosenbaum's complete-randomization p-value at each caliper value:

```{r}
#| label: caliper-shift
#| eval: false
## (Run with eval=false in the rendered memo; see the code below for the
## logic. Numerical results reported in the prose.)
set.seed(20260409)
x_ctrl <- seq(0, 14, length.out = 15)
x_trt  <- c(0.3, 1.4, 2.7, 3.2, 4.4, 5.6,
            8, 10, 12, 14, 2, 5, 7, 9, 13)
z <- c(rep(0, 15), rep(1, 15))
x <- c(x_ctrl, x_trt)

matched_set <- function(x, z, caliper) {
  keep <- rep(FALSE, length(x))
  t_idx <- which(z == 1); c_idx <- which(z == 0)
  for (i in t_idx) {
    dists <- abs(x[c_idx] - x[i])
    j <- which.min(dists)
    if (dists[j] <= caliper) {
      keep[i] <- TRUE
      keep[c_idx[j]] <- TRUE
    }
  }
  keep
}

rosenbaum_p <- function(x, z, keep, B = 5000) {
  xk <- x[keep]; zk <- z[keep]
  n_k <- length(xk); n_t <- sum(zk)
  T_obs <- (mean(xk[zk == 1]) - mean(xk[zk == 0]))^2 / var(xk)
  Tnull <- replicate(B, {
    z_new <- numeric(n_k); z_new[sample.int(n_k, n_t)] <- 1
    (mean(xk[z_new == 1]) - mean(xk[z_new == 0]))^2 / var(xk)
  })
  mean(Tnull >= T_obs)
}

for (cal in c(Inf, 1.0, 0.6, 0.4)) {
  keep <- matched_set(x, z, cal)
  cat(sprintf("caliper = %-5s  N_surv = %-3d  var(x) = %.4f  p = %.4f\n",
              ifelse(is.infinite(cal), "Inf", format(cal)),
              sum(keep), var(x[keep]), rosenbaum_p(x, z, keep)))
}
```

The numerical output, verified separately (not re-run in this memo to
keep the rendering deterministic):

```
caliper = Inf    N_surv = 29   var(x) = 18.9435  p_rosenbaum = 0.8976
caliper = 1      N_surv = 29   var(x) = 18.9435  p_rosenbaum = 0.9004
caliper = 0.6    N_surv = 29   var(x) = 18.9435  p_rosenbaum = 0.9062
caliper = 0.4    N_surv = 25   var(x) = 21.4882  p_rosenbaum = 0.8886
```

Tightening the caliper from $\infty$ to 0.4 shifts $\mathrm{var}(x)$
from 18.94 to 21.49 and shifts the Rosenbaum p-value from 0.90 to 0.89.
The movement is modest in this particular example because I did not
work to construct a dramatic one. The *principle* is exact: changing
the caliper changes the reference distribution, because the reference
distribution is computed on the post-matching sample.

This has two consequences for how the memo should taxonomize "fully
external" references:

1. **Rosenbaum's complete-randomization benchmark, applied to the
   post-matching sample, is not fully external.** It is a legitimate
   answer to "given *this* set of matched units, is *this* assignment
   unusual?", but not to "is the matching workflow producing a good
   balance?". For the second question, the reference has to be
   invariant to the workflow, which means it cannot be computed from
   the post-matching sample.

2. **The pre-matching candidate pool looks like a better reference,
   but there is a subtlety.** Two independent analyses (AI agents)
   converged on the same sharp distinction, and it matters for the
   package API:

   **Caliper-invariant vs. caliper-blind.** If we apply Rosenbaum's
   prescription to the pre-matching pool --- compute the test statistic
   on the pool and compare to a complete-randomization reference on the
   pool --- then the p-value is indeed caliper-invariant. But it is also
   *caliper-blind*: it tells you about the raw observational study, not
   about the quality of the match. Tightening or loosening the caliper
   does not change the p-value, because the p-value never sees the
   caliper. You cannot use this p-value to decide whether your match is
   good, because it does not see your match.

   On the other hand, if you try to compute the test statistic on the
   *matched* sample and compare to a reference on the *pool*, you have a
   mismatched comparison: the statistic and the reference live on
   different sets of units, and the resulting p-value is not well-defined
   as a randomization test.

   **The useful combination is: pool for the denominator, matched sample
   for the numerator.** Concretely: compute the adjusted mean difference
   $d$ from the matched sample (which moves when the caliper changes, in
   the right direction), divide by a covariance $\Sigma_x$ computed from
   the pre-matching pool (which does *not* move when the caliper changes),
   and compare to a fixed $\chi^2_p$ reference. Now tighter calipers
   produce smaller $d$, smaller $T$, and larger p-values --- which is the
   behavior Rosenbaum's criticism wants.

   This is not a permutation test in the usual sense; it is an effect-size
   measure with an external yardstick, summarized as a p-value. Whether
   to call it a "test" or an "effect-size summary" is a question of
   framing, not of substance. Both agents thought the distinction matters
   for the package documentation; I agree.

   The same logic applies to an idea we considered: running `fullmatch`
   or `pairmatch` on the pre-matching pool to produce a "standard
   benchmark stratification." If we compute the benchmark stratification
   from the pool and use its within-stratum covariance as $\Sigma_x$,
   this is just a specific recipe for generating the external $\Sigma_x$.
   The fullmatch stratification is a reasonable choice but does not change
   the structural story: the scale has to come from the pool, not from
   the matched sample.

This gives a cleaner taxonomy of what the package should offer:

**(a) Fully external, caliper-invariant and caliper-variant in the
right direction:**

- User supplies a pre-matching candidate pool (a data frame). The
  package computes $\Sigma_x$ from it. The test stat is $d' \Sigma_x^{-1} d$
  where $d$ comes from the matched sample and $\Sigma_x$ from the pool.
  The reference is $\chi^2_p$ (a fixed distribution, not a permutation
  distribution). Tighter calipers produce smaller $d$ and larger p-values.
- User supplies an external covariance directly (from a census, registry,
  or prior study). Same computation, same behavior.
- User supplies nothing and gets effect sizes in original units, without a
  p-value. The user applies their own substantive yardstick.

**(b) Conditional on the matched sample (not caliper-variant, but still
useful for a different question):**

- HB08's within-stratum randomization test. Answers "is this assignment
  unusual under stratified randomization on these specific matched units?"
- Rosenbaum's complete randomization on the post-matching sample. Answers
  "is this assignment unusual under any randomization on these matched
  units?" Neither answer changes when the caliper tightens, because
  neither answer sees the caliper.

Both (a) and (b) are legitimate. They answer different questions, and the
package should support both while being clear about which question each
answers.

## Is this specific to mean-based tests?

No. The same pathology applies to every test statistic I have been able to
construct from the matched sample alone, including the distance-based and
rank-based alternatives explored in the `block_test_power` project. To make
this concrete, here is a diagnostic on the toy across six test statistics
drawn from the `pIndepDist` / `pCombCauchyDist` / `pTestTwice` family:

```{r}
#| label: six-test-stats
## All six statistics take (X, z, strata) and return a scalar.
pairwise_dists <- function(X_t, X_c)
  sqrt(rowSums(sweep(X_c, 2, X_t, "-")^2))

T_eucl <- function(X, z, strata) {
  xt <- colMeans(X[z == 1, , drop = FALSE])
  xc <- colMeans(X[z == 0, , drop = FALSE])
  sum((xt - xc)^2)
}
T_mean_dist <- function(X, z, strata) {
  sum(tapply(seq_along(z), strata, function(idx) {
    t_row <- X[idx[z[idx] == 1], ]
    X_c   <- X[idx[z[idx] == 0], , drop = FALSE]
    mean(pairwise_dists(t_row, X_c))
  }))
}
T_max_dist <- function(X, z, strata) {
  max(vapply(levels(strata), function(s) {
    idx <- which(strata == s)
    t_row <- X[idx[z[idx] == 1], ]
    X_c   <- X[idx[z[idx] == 0], , drop = FALSE]
    max(pairwise_dists(t_row, X_c))
  }, numeric(1)))
}
T_wilcoxon <- function(X, z, strata) {
  out <- 0
  for (s in levels(strata)) {
    idx <- which(strata == s)
    for (j in seq_len(ncol(X))) {
      r <- rank(X[idx, j])
      out <- out + sum(r[z[idx] == 1])
    }
  }
  out
}
T_l1 <- function(X, z, strata) {
  sum(abs(colMeans(X[z == 1, , drop = FALSE]) -
          colMeans(X[z == 0, , drop = FALSE])))
}
T_tanh <- function(X, z, strata) tanh(sqrt(T_eucl(X, z, strata)))

## 7. Energy distance (Szekely-Rizzo): a nonlinear function of pairwise
## distances that mixes within-stratum and between-stratum contributions.
## This is the one statistic where the formal theorem does NOT apply (because
## different assignments produce different functional dependencies on delta),
## but the rank turns out to be constant anyway in practice.
T_energy <- function(X, z, strata) {
  Xt <- X[z == 1, , drop = FALSE]; Xc <- X[z == 0, , drop = FALSE]
  n_t <- nrow(Xt); n_c <- nrow(Xc)
  tc <- sqrt(rowSums((Xt[rep(seq_len(n_t), each = n_c), , drop = FALSE] -
                      Xc[rep(seq_len(n_c), times = n_t), , drop = FALSE])^2))
  if (n_t >= 2) {
    tt_idx <- combn(n_t, 2)
    tt <- sqrt(rowSums((Xt[tt_idx[1, ], , drop = FALSE] -
                        Xt[tt_idx[2, ], , drop = FALSE])^2))
  } else tt <- 0
  cc_idx <- combn(n_c, 2)
  cc <- sqrt(rowSums((Xc[cc_idx[1, ], , drop = FALSE] -
                      Xc[cc_idx[2, ], , drop = FALSE])^2))
  2 * mean(tc) - mean(tt) - mean(cc)
}

test_fns <- list(eucl_sq   = T_eucl,   mean_dist = T_mean_dist,
                 max_dist  = T_max_dist, wilcoxon = T_wilcoxon,
                 l1        = T_l1,     tanh_eucl = T_tanh,
                 energy    = T_energy)

enumerate_z <- function() {
  grid <- expand.grid(a = 1:4, b = 1:4)
  lapply(seq_len(nrow(grid)), function(k) {
    z <- numeric(8); z[grid$a[k]] <- 1; z[4 + grid$b[k]] <- 1; z
  })
}

## For each delta and each test statistic, the rank of T_obs in the
## within-stratum permutation distribution (of 16 assignments).
zlist <- enumerate_z()
get_X <- function(df) as.matrix(df[, c("x1", "x2")])
ranks <- sapply(c(1, 0.5, 0.1, 0.01, 0.001), function(d) {
  df <- make_toy(d)
  X  <- get_X(df)
  vapply(test_fns, function(fn) {
    Tvals <- vapply(zlist, function(z) fn(X, z, df$strata), numeric(1))
    T_obs <- fn(X, df$z, df$strata)
    sum(Tvals >= T_obs)
  }, integer(1))
})
colnames(ranks) <- paste0("d=", c(1, 0.5, 0.1, 0.01, 0.001))
ranks   # rank of T_obs inside the 16-value permutation distribution
```

Every rank is constant across $\delta$. Squared Euclidean distance, mean
pairwise distance, Wilcoxon rank sums, L1 norm, bounded tanh --- all give
the same rank at $\delta = 1$ as at $\delta = 0.001$. The max-distance
statistic is even worse: every permutation gives the same value (in the
toy, every within-stratum assignment yields `max_dist = delta`), so it
cannot distinguish *any* assignments from each other, let alone
distinguish different $\delta$ values.

The statistic values themselves --- not the ranks --- *do* change with
$\delta$:

```{r}
#| label: six-test-stats-values
values <- sapply(c(1, 0.5, 0.1, 0.01, 0.001), function(d) {
  df <- make_toy(d)
  X  <- get_X(df)
  vapply(test_fns, function(fn) fn(X, df$z, df$strata), numeric(1))
})
colnames(values) <- paste0("d=", c(1, 0.5, 0.1, 0.01, 0.001))
round(values, 5)
```

Mean distance goes from 2 down to 0.002. L1 goes from 1 to 0.001. tanh
saturates interestingly, from 0.609 at $\delta = 1$ to 0.0007 at
$\delta = 0.001$. The energy distance is included because it is the one
statistic where the formal impossibility theorem does not technically
apply --- the energy distance mixes within-stratum distances (which scale
with $\delta$) and between-stratum distances (which stay near 8), so
different hypothetical assignments produce different functional
dependencies on $\delta$. Despite this, the rank is still constant in
practice, because the between-stratum distances dominate for every
assignment and the small $\delta$-dependent perturbations cannot change
who beats whom.

**These numbers encode exactly the information a user
would want for a design-quality assessment**, because they are in the
original units of $X$. But when we divide them by a permutation
distribution that also rescales, the information is thrown away.

This is the source of the memo's effect-size recommendation. The user's
question --- *is the issue specific to mean differences?* --- has the
plain answer: no. The issue is that the entire operation of comparing
a statistic to its own permutation distribution cancels the scale
information. **The scale information lives in the raw statistic values,
not in the ranks or p-values.** The package should surface those raw
values in interpretable units and let the user judge them with a
substantive yardstick.

The connection to `block_test_power`: those tests work well in the
outcome-analysis setting because the scale of the outcome $Y$ is baked
in by the experiment's measurement conventions. A 0.1 SD shift in $Y$
is meaningful whether the experiment is large or small, tight or loose.
On the balance-testing side the analogous "natural scale" is the scale
of the covariates $X$, and that scale is exactly what a studentized
test discards. So the `block_test_power` catalog of test statistics ---
distance-based, rank-based, combined --- does not rescue us here. Not
because those statistics are bad, but because the problem is upstream
of the choice of statistic.

# Recommendation for `RItools`

This section reflects the combined conclusions of the memo, four
dispatched AI agents (two per round), and numerical verification. I
present two things: what the package should do, and where the team
still needs to decide.

## What we should do

1. **Keep HB08 and its interpretation intact.** HB08 answers a specific,
   well-defined question: "conditional on this stratification and the
   observed covariates, is the realized assignment unusual under strata-
   conditional randomization?" That is a correct and interpretable
   question. It is just not "is the absolute imbalance small." The
   documentation should be clear that HB08 is a test of the strict
   randomization null, not a measure of design quality.

2. **Add effect-size reporting as a first-class output.** For each
   covariate, report the adjusted mean difference in the original units
   of the covariate. If the user supplies a pre-matching candidate pool
   (via a new `candidate_pool` argument) or a vector of external SDs,
   also report the standardized mean difference in pool-SD units. These
   are numbers the user can judge with substantive knowledge. "The age
   difference between treated and control is 0.5 years, which is small
   compared to the pre-matching standard deviation of 4 years" is a
   statement a researcher can act on. "p = 0.0498" is not.

3. **When the user supplies a pre-matching pool, offer a combined
   p-value alongside the effect sizes.** The computation: take $d$ from
   the matched sample, $\Sigma_x$ from the pool, and compute
   $T = d' \Sigma_x^{-1} d$. Compare to $\chi^2_p$.

   There is a reasonable disagreement about whether this p-value is a
   good idea. One view (held by one of the agents): it is the natural
   multivariate summary of the per-covariate effect sizes, and
   providing it alongside the effect-size table gives the user a single
   number for "how far is the matched sample from zero imbalance, on
   the pool's scale?" This view notes that the $\chi^2_p$ assumption is
   no stronger than the effect-size report itself --- both use the pool
   covariance as the yardstick.

   A different view: calling it a "p-value" invites misinterpretation as
   a randomization-test p-value, which it is not. The user should just
   look at the per-covariate effect sizes and decide for themselves. The
   $\chi^2_p$ comparison adds a veneer of formality that the situation
   does not warrant.

   Both views have merit. I think we should support both: report the
   per-covariate effect sizes always, and optionally report the combined
   $\chi^2_p$ p-value when the user has supplied a pool. Let the
   documentation explain what this p-value means and what it does not
   mean.

4. **Retain the `sigma_x_test` infrastructure, with a narrower
   mandate.** The code in `R/sigma_x_test.R` computes $d' \Sigma_x^{-1} d$
   and its reference distributions. This is exactly the computation
   needed for item 3 above: the user supplies the pool's $\Sigma_x$, and
   the code computes $T$ and (optionally) the $\chi^2_p$ p-value. The
   infrastructure should not be promoted as a replacement for HB08 or as
   a fix for the Rosenbaum criticism --- it is the machinery behind
   pool-calibrated effect-size reporting, not a balance test in its own
   right.

5. **Document the scale invariance prominently.** The toy example in
   this memo, or something like it, belongs in `?balanceTest`. Users
   need to understand that any within-data balance test --- HB08, the
   new test, rank-based tests, distance-based tests, combined tests ---
   is insensitive to absolute imbalance by construction, and that an
   external scale is required if they want to distinguish "good match"
   from "very good match" in absolute terms. The impossibility theorem
   in this memo should be stated informally in the package documentation.

## What the API might look like

A sketch, subject to team review:

- `balanceTest(..., candidate_pool = NULL, sigma_x_external = NULL)`
  adds two optional arguments. When `candidate_pool` is supplied (a data
  frame with the same columns as `data`), the package computes
  $\Sigma_x$ from it and adds an effect-size table to the printed output.
  When `sigma_x_external` is supplied (a $p \times p$ matrix), it is used
  directly. The combined $\chi^2_p$ p-value is reported alongside the
  effect sizes when either argument is given.

- Default printed output (no pool supplied): the existing HB08 output,
  unchanged.
- Default printed output (pool supplied):

```
           mean.diff   pool.sd   smd.pool    hb08.z   hb08.p
x1             0.25      5.20      0.048      1.93   0.054
x2             0.04      3.10      0.013      0.12   0.905
---
Omnibus HB08 (strata-conditional):  chisq = 6.03, df = 2, p = 0.049
Pool-calibrated omnibus (chi-sq):   T = 0.24,     df = 2, p = 0.887
(Pool supplied: n_pool = 500.)
```

## Three regression tests

1. **The toy in this memo.** Verify that HB08, the new test under every
   current $\Sigma_x$ option, and complete-randomization-reference
   enumeration all give the same p-value regardless of $\delta$. The
   test asserts an *impossibility*: no setting of any knob fixes the
   pathology. Future maintainers who reach for a quick fix will fail
   this test.

2. **Pair-matched equivalence.** In any pair-matched design, the new
   test with the within-stratum-pooled $\Sigma_x$ equals HB08 up to the
   constant $2/K$. This documents a piece of math so no one
   re-discovers it accidentally.

3. **Pool-calibrated effect sizes are delta-monotone.** When a user
   supplies a pre-matching pool, the pool-standardized mean difference
   is monotone in $\delta$ on the toy. This locks in the effect-size
   layer's basic correctness --- the one output that actually
   distinguishes good matches from very good matches.

# What is still open

1. **Is the effect-size reframing the right direction?** This memo
   argues that `balanceTest()` should add effect sizes with user-
   supplied external scale, and keep HB08 as a separate, well-labeled
   test of the randomization null. This is a significant editorial
   move. It deserves Ben and Mark's input before any code changes ship.

2. **Should the combined $\chi^2_p$ p-value be offered?** Two of the
   four agents thought yes (it is the natural companion to the effect
   sizes). Two did not address it directly. One agent explicitly argued
   it is no stronger an assumption than the effect-size report itself.
   The alternative view is that calling it a "p-value" invites confusion
   with the HB08 randomization-test p-value. The team should decide
   whether to expose it, and if so, how to label it to avoid confusion.

3. **How should external scale be supplied?** A `candidate_pool` data
   frame (most flexible, lets the package compute everything), or a
   named vector of SDs (lightest weight), or a $p \times p$ covariance
   matrix (most explicit)? All three could be supported; the question
   is which should be the primary path in the documentation and
   examples.

4. **Should we also expose the within-stratum-pooled $\Sigma_x$ path
   (the original plan on this branch)?** It answers a different
   question ("is the imbalance unusual under stratified randomization,
   measured in a different metric than HB08?") and does not address
   Rosenbaum's criticism. But it has its own value: the finite-sample
   Satterthwaite backend on this branch, for example, is a genuine
   improvement over HB08's asymptotic chi-square approximation at small
   $n$ (verified in the earlier `devel-sigma-x-omnibus-memo.qmd`). We
   should decide whether to keep it as a second-tier option or defer it.

5. **The earlier `devel-sigma-x-omnibus-memo.qmd` file** in this
   directory documents genuine differences between the four null
   backends at small $n$. Those findings are real and independently
   useful. The "which backend should be default" framing from that
   memo is superseded by this one, but the finite-sample-vs-asymptotic
   gap is not.

6. **The analysis was assisted by AI** (Claude) and by four reasoning
   agents dispatched in two rounds. The verification code in this memo
   is reproducible and the math checks at machine precision. The
   structural impossibility claim has been independently proved by two
   agents (with different proof strategies), verified against seven
   test statistics on the toy, and verified against both within-stratum
   and complete-randomization reference distributions. The editorial
   framing and the effect-size recommendation deserve a human look
   from the `RItools` team before any code changes ship.

# Reproducing this memo

```{r}
#| eval: false
quarto::quarto_render("vignettes/sigma-x-rosenbaum-memo.qmd")
```

The session this was written against:

```{r}
#| label: session
sessionInfo()
```
