---
title: "Balance collapse, screening, and the d^2/ACAT switch"
subtitle: "A simulation memo for the RItools team"
author: "Jake Bowers (with Claude)"
date: 2026-06-17
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    embed-resources: true
    df-print: kable
  pdf:
    toc: true
    toc-depth: 3
    df-print: kable
    geometry: margin=1in
    code-overflow: wrap
    fig-width: 6.5
    fig-height: 5
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>",
                      fig.width = 9, fig.height = 7, dpi = 130)
options(keep.source = TRUE)
## The reusable code -- data generator, screens, ESS, and the omnibus runners --
## lives in sim-helpers.R and is sourced here.  The experiment code for each
## result is shown inline below.
source("sim-helpers.R")
NREP <- 1000L
```

These simulations settle two practical questions about `balanceTest()`: (1) what
does the d^2 omnibus actually respond to, so we know what to tell users it means;
and (2) are the two proposed changes -- a screening step for near-exactly-matched
covariates, and an automatic switch to the Cauchy (ACAT) omnibus -- worth
building, and for what reason.

The reusable functions are in `vignettes/sim-helpers.R` (sourced above); every
experiment's own code is shown in the chunk that produces its result.

::: {.callout-note}
## Framing update (2026-06-17) and status

Two things have moved since the first draft, and the next revision will restructure
this memo around them.

First, the synthesis. The recommended reading of the omnibus is now SIZE x
PRECISION: the chi-square is a fixed-scale magnitude (the standardized
differences) times a precision that grows as the match tightens, so a small
omnibus p with small standardized differences is the precision, not the
imbalance. The full argument, and the result that no comparison to a
block-randomized standard on these same strata can track the size of imbalance
(a theorem for the collapse, a strongly supported conjecture for the no-escape
claim, after a nine-way adversarial search), are written up in
`comparing-design-to-randomized-standard-memo.md` and
`impossibility-pressure-test-memo.md`. Read those for the current position; the
simulations below are the supporting evidence.

Second, the two proposed changes are now built into `balanceTest()`. The screen
is a RELATIVE within-stratum-variance tolerance: a covariate that retains a
negligible fraction of its variance within strata is dropped from the omnibus
with a message naming it, and the omnibus abstains gracefully (df 0 / NA p)
rather than erroring when every covariate is near-exactly matched. It is a
numerical-tolerance and honest-abstention step, not a fix for the collapse. The
Cauchy combination (ACAT) omnibus is available via `cauchy.combination = TRUE`.

The provenance of the 0.25-SD rule has moved to its own file
(`balance-threshold-provenance.md`); Jake's current view is to prefer a
substantive, design-stage tolerance (for example, a maximum age difference in
years) over a universal standardized cutoff. Several Study-1 / Study-2
subsections below (the p-vs-ESS jaggedness, the gaming and direction-screen
demonstrations, the power study, and the provenance) are slated to move out in
the restructure; they are kept here for now as the record of what was checked.
:::

# The data-generating process and the knobs

Every experiment uses matched **pairs** (one treated, one control per set),
because pairs make the algebra transparent and because the matched-pair case is
where the collapse is sharpest. There are $k$ covariates, $x_1,\dots,x_k$, and
they are **correlated** across columns. For each pair $s$ and covariate $j$:

- a pair **location** $m_{sj}$ -- the BETWEEN-pair variation, the part of
  covariate $j$ that stratification removes. It plays no role in any test below,
  and it is drawn independently of the within-pair difference.
- a treated-minus-control **difference** $D_{sj} = \text{noise}_{sj} + \mu_j$,
  where $\text{noise}_{sj}$ is symmetric (random sign, pair to pair) with marginal
  SD $\sigma_{c}[j]$, and $\mu_j$ is a constant added to every pair on covariate
  $j$.
- the treated unit's value is $m_{sj} + D_{sj}/2$, the control's is
  $m_{sj} - D_{sj}/2$.

`mu` and `sigma_c` are length-$k$ **vectors**: `mu[j]` is the systematic tilt on
covariate $j$, `sigma_c[j]` is its within-pair noise SD. The covariates are not
exactly standardized: each location has unit variance and each within-pair noise
has marginal SD `sigma_c[j]` (= 1 everywhere except near-exactly-matched
covariates, where it is ~0), so each covariate sits on a roughly unit scale and
`mu[j]` is approximately in SD units. The quantity the test reads is
$\mu_j/\sigma_c[j]$ (the "standardized bias"). Correlation is moderate
equicorrelation (every pair of covariates 0.5), applied to both the locations and
the within-pair differences, with column scaling that preserves each marginal SD;
`rho` is one tunable argument (`rho = 0` recovers independent covariates).

The data generator:

```{r show-make-pairs}
make_pairs
```

Two short helpers used inline below: `scale_within()` applies the size knob $t$
(scale the whole within-pair difference by $t$, direction fixed), and `smd()` is
the Love-plot standardized mean difference (mean difference over pooled SD).

```{r inline-helpers}
scale_within <- function(df, t) {
  for (cn in covariate_names(df)) {
    xt <- df[[cn]]; pm <- ave(xt, df$pair)
    df[[cn]] <- pm + (xt - pm) * t          # scale within-pair part, keep pair mean
  }
  df
}
smd <- function(df, cn) {
  x <- df[[cn]]; g1 <- df$z == 1; g0 <- df$z == 0
  (mean(x[g1]) - mean(x[g0])) / sqrt((var(x[g1]) + var(x[g0])) / 2)   # pooled-SD SMD
}
```

## What the test responds to

The d^2 omnibus is $d^2 = \bar d' V_d^{-1} \bar d$, with $\bar d$ the average
treated-minus-control difference and $V_d$ its covariance under within-pair
re-randomization. With this DGP, $\bar d \approx \mu$ (the noise averages out) and
$V_d \approx (1/S)\,(\text{within-pair covariance})$, so for a single covariate

$$d^2 \;\approx\; S \,\Big(\tfrac{\mu}{\sigma_c}\Big)^2 .$$

The omnibus responds to the systematic tilt **measured in units of the noise**
($\mu/\sigma_c$) times the sample size $S$. Under the null it is central
chi-square on $\mathrm{rank}(V_d)$ df; under a tilt, noncentral chi-square with
$\lambda \approx S\,(\mu/\sigma_c)^2$. With correlated covariates the multivariate
version is $\lambda = \mu' V_d^{-1}\mu$, and the alignment of $\mu$ with $V_d$'s
eigenvectors matters -- which is what makes B4 flip.

Three consequences (= A1, A2, A3): scale $\mu$ and $\sigma_c$ together and the
ratio is unchanged, so d^2 does not move (A1, the collapse); shrink $\mu$ with
$\sigma_c$ fixed and d^2 falls (A2); hold the ratio and grow $S$ and power rises
(A3).

## Two standardizations -- why this differs from a Love plot

The Love-plot standardized mean difference divides by the **pooled SD**, which is
dominated by between-pair variation that the size knob $t$ leaves alone -- so the
SMD scales with $t$ and registers when groups get more or less different. d^2
divides by the **within-pair null SD** (the $\sigma_c$ that $t$ shrinks), so the
$t$ cancels. The Love plot sees the size of imbalance; d^2 does not. They agree
only when $\sigma_c$ is held fixed (which is what A2 does).

# Study 1: what the omnibus does and does not respond to

## A1 -- the collapse: d^2 ignores the size of the imbalance

One fixed data set with a tilt on $x_1$; increase or decrease the size of the
whole within-pair imbalance by $t$ (direction fixed) over five orders of
magnitude.

```{r A1}
set.seed(1)
base_df <- make_pairs(S = 100, sigma_c = rep(1, 4), mu = c(0.5, 0, 0, 0))
a1 <- do.call(rbind, lapply(c(100, 10, 1, 0.1, 0.01, 0.001), function(t) {
  r <- run_omnibus(scale_within(base_df, t))
  data.frame(t = t, chisq = round(r$chisq, 3), df = r$df,
             p_d2 = signif(r$p_d2, 3), p_acat = signif(r$p_acat, 3))
}))
a1
```

A thousand-fold change in how different the treated and control units actually
are produces no change in the statistic or its p-value. This is the collapse,
exact: d^2 is homogeneous of degree zero in the imbalance (numerator scales by
$t$, $V_d$ by $t^2$, $t\cdot t/t^2 = 1$). A Love plot would show balance improving
as $t$ falls; d^2 does not.

Two distinct limits show up, and only one is escapable. **ACAT is invariant too**
(`p_acat` flat): it combines per-covariate p-values, each itself a size-blind
within-pair test, so the combination is size-blind. Nothing on the within-pair
reference sees the size of imbalance. The **other** limit, which ACAT does escape,
is the cap on how many covariates d^2 can test: d^2 inverts $V_d$, whose rank
cannot exceed the within-block residual df; past that d^2 freezes (p near 0.45).
ACAT never inverts anything (shown in B4).

## A2 -- the test IS informative about systematic confounding

Hold the within-pair noise fixed (`sigma_c = 1`) and shrink only the systematic
tilt `mu` on one covariate from 0.5 to 0.

```{r A2}
set.seed(20260617)
mu_grid <- c(0.5, 0.3, 0.2, 0.1, 0.05, 0)
a2 <- do.call(rbind, lapply(mu_grid, function(mu1) {
  ps <- replicate(NREP, run_omnibus(make_pairs(100, rep(1, 3), c(mu1, 0, 0)))$p_d2)
  data.frame(mu = mu1, rejection_rate = rejection_rate(ps),
             median_p = round(median(ps, na.rm = TRUE), 4))
}))
a2
```

Unlike A1, the test moves, and correctly. A large tilt is detected almost always;
as the tilt vanishes the rejection rate falls to the nominal 0.05 and the
p-values become uniform (median ~0.5). So the omnibus is a sound test of the
systematic, direction-carrying part of imbalance. A1 and A2 together: it reads
$\mu/\sigma_c$. A1 moved both (ratio fixed -> no response); A2 moved $\mu$ alone
(ratio falls -> response). "Collapse" is specific to the size dimension.

## A3 -- more data detects a fixed bias more often

Hold the standardized bias fixed at $\mu/\sigma_c = 0.15$ and grow the study from
25 to 400 pairs (ESS = pairs).

```{r A3}
set.seed(20260617)
a3 <- do.call(rbind, lapply(c(25, 50, 100, 200, 400), function(S) {
  ps <- replicate(NREP, run_omnibus(make_pairs(S, rep(1, 3), c(0.15, 0, 0)))$p_d2)
  data.frame(ESS = S, rejection_rate = rejection_rate(ps))
}))
a3
```

With the same real bias throughout, power climbs from ~0.10 to ~0.90: ordinary
power, $\lambda \approx S\,(\mu/\sigma_c)^2$. The flip side is the warning: because
power against any fixed nonzero bias keeps rising with ESS, a large enough matched
sample rejects on a bias too small to matter. That "balance tests reject in large
samples" problem is a second reason -- on top of the collapse -- to report a
fixed-scale magnitude rather than read the design off the omnibus p.

## A4 -- why a p-value-vs-ESS plot is jagged

One fixed population of 200 confounded pairs (a modest tilt on $x_1$). Prune
pairs worst-offenders-first (most aligned with the bias), recompute the omnibus
at each step, and track the p-value and the statistic against ESS.

```{r A4}
set.seed(7)
S0 <- 200
df_full <- make_pairs(S = S0, sigma_c = rep(1, 4), mu = c(0.10, 0, 0, 0))
align <- tapply(df_full$x1 * ifelse(df_full$z == 1, 1, -1), df_full$pair, sum)
prune_order <- order(align, decreasing = TRUE)            # drop most-aligned first
lvl <- levels(df_full$pair)
path <- do.call(rbind, lapply(seq(S0, 30, by = -5), function(nkeep) {
  keep <- prune_order[(S0 - nkeep + 1):S0]                # keep the least-aligned nkeep
  sub  <- droplevels(df_full[df_full$pair %in% lvl[keep], ])
  r <- run_omnibus(sub)
  std_bias <- abs(mean(tapply(sub$x1 * ifelse(sub$z == 1, 1, -1), sub$pair, sum)))
  data.frame(ess = effective_sample_size(sub), df = r$df, chisq = r$chisq,
             p_d2 = r$p_d2, p_acat = r$p_acat, std_bias = std_bias)
}))
## representative rows
path[path$ess %in% c(200, 195, 185, 180, 100, 30),
     c("ess", "chisq", "p_d2", "p_acat", "std_bias")]
```

As ESS falls the p-value is non-monotone: it ticks up near the full sample (where
pruning removes the targeted tilt) and then crashes. The honest driver is the
**statistic itself** -- d^2 is a hump-shaped function of ESS (low at the full
sample, peaking at intermediate ESS as pruning overshoots and induces imbalance,
then falling as the sample shrinks). Since $p = \mathrm{pchisq}(d^2, \mathrm{df},
\text{lower}=F)$, the p-value is the monotone transform of that hump.

```{r A4-figure}
#| fig-cap: "Study 1. A2: p tracks systematic bias. A3: power grows with ESS. A4: p vs ESS is non-monotone (the p-axis crushes it near 0); the de-compressed d^2 statistic shows the hump that drives it."
op <- par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(a2$mu, a2$rejection_rate, type = "b", pch = 19, ylim = c(0, 1),
     xlab = "systematic bias mu (sigma_c = 1)", ylab = "rejection rate (alpha=.05)",
     main = "A2: p tracks systematic bias")
abline(h = 0.05, lty = 2, col = "grey50")
plot(a3$ESS, a3$rejection_rate, type = "b", pch = 19, ylim = c(0, 1),
     xlab = "effective sample size (= pairs)", ylab = "rejection rate (alpha=.05)",
     main = "A3: ESS buys power")
abline(h = 0.05, lty = 2, col = "grey50")
plot(path$ess, path$p_d2, type = "b", pch = 19, ylim = c(0, 1),
     xlab = "effective sample size", ylab = "omnibus p-value",
     main = "A4: p vs ESS (non-monotone)")
lines(path$ess, path$p_acat, type = "b", pch = 1, col = "red")
legend("topleft", c("d^2", "ACAT"), pch = c(19, 1), col = c("black", "red"), bty = "n")
plot(path$ess, path$chisq, type = "b", pch = 19, col = "blue",
     xlab = "effective sample size", ylab = "omnibus d^2 statistic (df=4)",
     main = "A4 driver: the d^2 statistic vs ESS")
par(op)
```

The payoff: $d^2 \approx S\,(\text{standardized imbalance})^2$, and along a pruning
path both factors move at once. You cannot read balance quality off a p-vs-ESS
curve -- its shape reflects which pairs you dropped as much as how balanced the
survivors are.

# Study 2: screening, and the d^2/ACAT switch

There are **two different kinds of screen**, and keeping them distinct is the
main lesson:

- A **design screen** asks "is this covariate (or direction) exactly matched?"
  The prototype is $w_j = (\text{within-stratum var})/(\text{total var})$ per
  covariate, which exact matching drives to ~0. It depends on covariate values
  and strata only -- not the treatment vector -- so it is ungameable (B5). The
  direction version (B6) generalizes it to linear combinations. It fixes the
  degenerate-direction pathology but is blind to the size of imbalance (B3).
- A **magnitude screen** asks "is the observed imbalance substantively large?"
  The prototype is the Love-plot rule: flag covariates with $|\mathrm{SMD}| \ge
  0.25$ (see "Provenance of the 0.25-SD rule" at the end of this section for the
  citation, which is *not* Rosenbaum-Rubin 1985 as often claimed). It uses the
  pooled-SD scale, so it does NOT collapse (B7); it is the constructive form of
  "lead with magnitude."

## B1 -- near-degenerate covariates cost d^2 power; the screen recovers it

Five covariates: $x_1,x_2$ near-exactly matched (`sigma_c = 1e-4`, no tilt), a
real confound on $x_3$, $x_4,x_5$ vary with no tilt. Detect $x_3$ three ways.

```{r B1}
set.seed(20260618)
b1 <- replicate(NREP, {
  df <- make_pairs(100, c(1e-4, 1e-4, 1, 1, 1), c(0, 0, 0.30, 0, 0))
  r_all <- run_omnibus(df)
  r_scr <- run_omnibus(df, covs = screen_survivors(df, tau = 0.01))
  c(p_all = r_all$p_d2, df_all = r_all$df,
    p_scr = r_scr$p_d2, df_scr = r_scr$df, p_acat = r_all$p_acat)
})
b1 <- as.data.frame(t(b1))
data.frame(
  method  = c("d^2 (all 5)", "d^2 (design-screened)", "ACAT (all 5)"),
  power   = round(c(rejection_rate(b1$p_all), rejection_rate(b1$p_scr),
                    rejection_rate(b1$p_acat)), 3),
  mean_df = round(c(mean(b1$df_all), mean(b1$df_scr), NA), 2))
```

The two dead covariates consume df and spread the chi-square over directions with
no signal; screening them out raises d^2 power. ACAT trails here because it spreads weight over five marginal
p-values, two of them pure noise: when the signal is concentrated and the
covariance is informative, the chi-square is the better omnibus.

## B1b -- graceful degradation when everything is (near-)exactly matched

```{r B1b}
set.seed(11)
df_ax <- make_pairs(100, rep(1e-9, 4), rep(0, 4))
r_all <- run_omnibus(df_ax)
surv  <- screen_survivors(df_ax, tau = 0.01)
cat(sprintf("d^2 (all): errored=%s, p=%s   <- hard stop() at R/utils.R:381\n",
            r_all$errored, format(r_all$p_d2)))
cat(sprintf("design screen: %d survivors -> %s\n", length(surv),
            if (length(surv) == 0) "ABSTAIN, no error" else "test survivors"))
cat(sprintf("ACAT: p=%s   <- NA, no error\n", format(r_all$p_acat)))
```

The current d^2 path throws when every covariate is degenerate. The screen turns
this into an honest non-result; ACAT returns NA cleanly. The package fix is
graceful degradation instead of erroring.

## B2 -- screening does not distort the type-I error rate

```{r B2}
set.seed(20260618)
b2 <- replicate(NREP, {
  df <- make_pairs(100, c(1e-4, 1e-4, 1, 1, 1), rep(0, 5))
  r_all <- run_omnibus(df)
  r_scr <- run_omnibus(df, covs = screen_survivors(df, tau = 0.01))
  c(p_all = r_all$p_d2, p_scr = r_scr$p_d2, p_acat = r_all$p_acat)
})
b2 <- as.data.frame(t(b2))
data.frame(method = c("d^2 (all)", "d^2 (screened)", "ACAT"),
           size_nominal_0.05 = round(c(rejection_rate(b2$p_all),
                                       rejection_rate(b2$p_scr),
                                       rejection_rate(b2$p_acat)), 3))
```

All three hold their size: the screen uses a design quantity, not the observed
imbalance, so it does not inflate false positives.

## B3 and B7 -- design vs magnitude screen, read against A1 on the SAME data

A1, B3, and B7 use the **identical** data set (the A1 base, `seed 1`) and the
**identical** $t$-sweep, so they read row for row. A1 is the omnibus with no
screen; B3 adds the design screen ($w_j$, `tau = 0.01`); B7 adds the magnitude
screen (flag $|\mathrm{SMD}| \ge 0.25$).

```{r B3-B7}
set.seed(1)
ab_base <- make_pairs(100, rep(1, 4), c(0.5, 0, 0, 0))
trio <- do.call(rbind, lapply(c(100, 10, 1, 0.5, 0.25, 0.1, 0.01), function(t) {
  df_t <- scale_within(ab_base, t)
  surv <- screen_survivors(df_t, tau = 0.01)          # design screen
  r    <- run_omnibus(df_t, covs = surv)
  smds <- vapply(covariate_names(df_t), function(cn) smd(df_t, cn), numeric(1))
  data.frame(
    t            = t,
    A1_omnibus_p = signif(run_omnibus(df_t)$p_d2, 3),
    B3_survivors = sprintf("%d/4", length(surv)),
    B3_screen_p  = if (is.na(r$p_d2)) "ABSTAIN" else signif(r$p_d2, 3),
    B7_SMD_x1    = round(smds[["x1"]], 3),
    B7_flagged   = sprintf("%d/4", sum(abs(smds) >= 0.25)))
}))
trio
```

**B3 (design screen is NOT a way around the collapse).** Where the screen keeps
covariates (large $t$), B3's p equals A1's collapsed value exactly. As $t$ shrinks
the within-stratum variance fraction falls below `tau` and the screen ABSTAINS --
it never returns a p that rewards the smaller imbalance.

**B7 (the Love-plot screen IS the non-collapsing one).** `SMD(x1)` tracks the size
of imbalance, crossing 0.25 between $t=1$ and $t=0.5$. At a well-matched design
($t=0.1$) the three verdicts diverge sharply: A1 still rejects (the collapse), B3
abstains, and B7 reports `SMD = 0.05 < 0.25` = balanced -- the substantively
correct answer. It escapes the collapse because the SMD uses the fixed pooled-SD
scale. Using the Love plot to screen is the constructive form of "lead with
magnitude"; the 0.25 threshold is an equivalence margin.

Two honest costs of the magnitude screen: (i) it is assignment-dependent (reads
the observed imbalance), so not ungameable like $w_j$; and (ii) you cannot screen
by observed SMD and then trust the within-pair omnibus p on the survivors --
selecting imbalanced covariates inflates that test. Use it as a magnitude/
equivalence **report**, not a pre-filter feeding the test.

## B6 -- the design screen should work on DIRECTIONS, not covariates

The per-covariate $w_j$ misses a degenerate **linear combination**. Here $x_1,x_2$
each vary within pairs (so each $w_j$ is moderate) but share the same within-pair
difference, so the $x_1 - x_2$ direction is near-exactly matched; a real confound
sits on $x_3$. The direction screen (`sv_screen_project` in `sim-helpers.R`)
generalizes $w_j$ to the generalized eigenvalues of within-vs-total covariance.

```{r B6-illustrate}
make_degen_combo <- function(S, eps = 5e-3) {
  df <- make_pairs(S, rep(1, 4), c(0, 0, 0.30, 0), rho = 0.3)
  dev1 <- df$x1 - ave(df$x1, df$pair)
  df$x2 <- ave(df$x2, df$pair) + dev1 + rnorm(nrow(df), sd = eps)  # x2 shares x1's within-pair diff
  df
}
set.seed(61)
d6 <- make_degen_combo(120)
round(within_var_fraction(d6), 4)                 # per-covariate w_j: all moderate
round(sv_screen_project(d6, tau = 0.01)$lambda, 5) # direction eigenvalues: one ~0
```

The per-covariate $w_j$ are all moderate (so that screen drops nothing) while the
direction screen finds one eigenvalue ~0 (the $x_1-x_2$ combination) and drops it.
Power to detect $x_3$:

```{r B6-power}
set.seed(62)
b6 <- replicate(NREP, {
  df <- make_degen_combo(120)
  c(p_all = run_omnibus(df)$p_d2,
    p_pc  = run_omnibus(df, covs = screen_survivors(df, tau = 0.01))$p_d2,
    p_sv  = run_sv_screen(df, tau = 0.01)$p_d2,
    p_acat = run_omnibus(df)$p_acat)
})
b6 <- as.data.frame(t(b6))
data.frame(method = c("unscreened d^2", "per-covariate screen",
                      "direction screen", "ACAT"),
           power  = round(c(rejection_rate(b6$p_all), rejection_rate(b6$p_pc),
                            rejection_rate(b6$p_sv), rejection_rate(b6$p_acat)), 3))
```

The per-covariate screen misses the degenerate combination (drops nothing, same
power as unscreened); the direction screen catches it and gives a sharper test.
The production screen should work on the singular values of $V_d$.

## B4 -- keep both omnibuses: under correlation ACAT is the more robust one

Three alternatives. "dense" = small tilt on all 8 covariates; "sparse" = one
strong tilt among 8; "highdim" = one strong tilt among 20 covariates with only 15
pairs (so #covariates exceeds $\mathrm{rank}(V_d)$).

```{r B4}
set.seed(20260618)
scen <- list(
  dense   = list(S = 100, sigma_c = rep(1, 8),  mu = rep(0.12, 8)),
  sparse  = list(S = 100, sigma_c = rep(1, 8),  mu = c(0.55, rep(0, 7))),
  highdim = list(S = 15,  sigma_c = rep(1, 20), mu = c(0.9, rep(0, 19))))
b4 <- do.call(rbind, lapply(names(scen), function(nm) {
  s <- scen[[nm]]
  out <- replicate(NREP, {
    r <- run_omnibus(make_pairs(s$S, s$sigma_c, s$mu)); c(r$p_d2, r$p_acat) })
  data.frame(scenario = nm, k = length(s$sigma_c), S = s$S,
             power_d2 = round(rejection_rate(out[1, ]), 3),
             power_acat = round(rejection_rate(out[2, ]), 3))
}))
b4
```

The **dense** case flips under correlation: with independent covariates d^2 won
(0.602 vs 0.456), but with positively correlated covariates ACAT wins and d^2's
power craters. This is correct: the dense tilt points in the all-covariates-equal
direction, the high-variance direction of the correlated $V_d$, and $d^2 = \bar d'
V_d^{-1}\bar d$ appropriately discounts imbalance there (it is less surprising).
ACAT ignores the covariance and keeps power -- the classic Hotelling-$T^2$ vs
marginal-combination tradeoff. The **sparse** case still slightly favors d^2; the
**highdim** case is decisive: d^2 has ~zero power (frozen near $p=0.45$; the
package warns "Degrees of freedom exceeds units less number of strata") while ACAT
keeps power. Net: ACAT is the more robust omnibus across correlation and
dimension.

```{r B-figure}
#| fig-width: 9
#| fig-height: 4
#| fig-cap: "Study 2. Left: d^2 vs ACAT power across dense/sparse/high-dim (under correlation, ACAT wins dense and is the only one with power in high dim). Right: detecting a real confound with two near-degenerate covariates present (design screening sharpens d^2)."
op <- par(mfrow = c(1, 2), mar = c(5, 4, 3, 1))
bp <- rbind(`d^2` = b4$power_d2, ACAT = b4$power_acat); colnames(bp) <- b4$scenario
barplot(bp, beside = TRUE, ylim = c(0, 1), legend.text = TRUE,
        args.legend = list(x = "topright", bty = "n"),
        ylab = "power (alpha=.05)", main = "B4: d^2 vs ACAT power")
b1bar <- c(`d^2 all` = rejection_rate(b1$p_all),
           `d^2 screen` = rejection_rate(b1$p_scr), `ACAT` = rejection_rate(b1$p_acat))
barplot(b1bar, ylim = c(0, 1), col = "grey80", ylab = "power to detect x3 (alpha=.05)",
        main = "B1: detect a confound with 2\nnear-degenerate covariates")
par(op)
```

## B5 -- the design screen cannot be gamed

```{r B5}
set.seed(5)
df5 <- make_pairs(80, c(1e-4, 1, 1), c(0, 0.3, 0))
w_obs <- within_var_fraction(df5)
w_reps <- replicate(50, {
  flip <- rep(rbinom(nlevels(df5$pair), 1, 0.5) == 1, each = 2)
  within_var_fraction(within(df5, z <- ifelse(flip, 1L - z, z)))
})
round(w_obs, 6)
cat(sprintf("max |w_j(re-randomized) - w_j(observed)| over 50 re-flips: %.2e\n",
            max(abs(w_reps - w_obs))))
```

$w_j$ does not depend on the treatment vector, so an analyst cannot manipulate
which covariates are screened by choosing an assignment. A screen based on the
observed z-statistic would not have this property.

## B8 -- how small can the magnitude-screen threshold be? (the floor)

The Love-plot screen (B7) tracks balance, but its threshold cannot be made
arbitrarily small. As $\tau \to 0$, the decision "balanced iff all
$|\mathrm{SMD}| < \tau$" converges to "balanced iff $\mathrm{SMD} = 0$" -- testing
the sharp null of exact balance, which is exactly where the collapse lives. So
there is a floor. On a BALANCED design ($\mu = 0$) the observed SMD is pure
sampling noise; the screen should rarely fire. The quantity that matters is
$\max|\mathrm{SMD}|$ across the $k$ covariates -- a balanced design is "flagged"
if any covariate trips the threshold (family-wise across $k$).

```{r B8}
maxabs_smd_balanced <- function(S, k, rho = 0.5) {
  df <- make_pairs(S, rep(1, k), rep(0, k), rho = rho)   # balanced: mu = 0
  max(abs(vapply(covariate_names(df), function(cn) smd(df, cn), numeric(1))))
}
set.seed(20260619)
b8 <- do.call(rbind, lapply(c(25, 50, 100, 200, 400), function(S) {
  m5  <- replicate(NREP, maxabs_smd_balanced(S, 5))
  m20 <- replicate(NREP, maxabs_smd_balanced(S, 20))
  data.frame(ESS = S,
             floor_k5        = round(as.numeric(quantile(m5,  0.95)), 3),
             flag_at_0.25_k5 = round(mean(m5  >= 0.25), 3),
             floor_k20        = round(as.numeric(quantile(m20, 0.95)), 3),
             flag_at_0.25_k20 = round(mean(m20 >= 0.25), 3))
}))
b8
```

`floor` is the threshold that yields a 5% false-flag rate (the 95th percentile of
$\max|\mathrm{SMD}|$ under balance); `flag_at_0.25` is the chance a balanced design
is flagged at $\tau = 0.25$.

```{r B8-figure}
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "B8. Magnitude-screen false-flag rate on a BALANCED design vs the threshold tau (k=5). As tau -> 0 the screen flags balanced designs ever more often (its version of the collapse). The curve shifts left as ESS grows (floor ~ 1/sqrt(ESS)). The 0.25 rule of thumb (dashed) is safe at large ESS but not small."
set.seed(20260619)
taus <- seq(0, 0.6, by = 0.02)
plot(NA, xlim = c(0, 0.6), ylim = c(0, 1),
     xlab = "magnitude-screen threshold tau (SMD)",
     ylab = "P(flag a balanced design)",
     main = "B8: SMD-screen false-flag rate vs threshold (k=5)")
cols <- c("25" = "red", "100" = "black", "400" = "blue")
for (S in c(25, 100, 400)) {
  m  <- replicate(NREP, maxabs_smd_balanced(S, 5))
  ff <- vapply(taus, function(tt) mean(m >= tt), numeric(1))
  lines(taus, ff, lwd = 2, col = cols[as.character(S)])
}
abline(v = 0.25, lty = 2, col = "grey40"); abline(h = 0.05, lty = 3, col = "grey60")
legend("topright", c("ESS=25", "ESS=100", "ESS=400", "tau=0.25"),
       col = c("red", "black", "blue", "grey40"),
       lty = c(1, 1, 1, 2), lwd = c(2, 2, 2, 1), bty = "n")
```

What it means. The floor behaves as

$$\tau_{\text{floor}} \;\approx\; \rho_{\text{match}}\,\frac{\sqrt{2\ln k}}{\sqrt{\text{ESS}}},
\qquad \rho_{\text{match}} = \frac{\text{rms within-pair difference}}{\text{pooled SD}},$$

so it (i) shrinks like $1/\sqrt{\text{ESS}}$ -- more data lets a smaller threshold
still work; (ii) grows like $\sqrt{\ln k}$ with the number of covariates, through
multiplicity (more covariates -> more chances for one to trip $\tau$ by noise);
and (iii) shrinks with matching tightness ($\rho_{\text{match}}$). These are loose
random pairs ($\rho_{\text{match}} \approx 0.9$, the worst case); a genuinely tight
match lowers the floor and makes 0.25 safer still.

The practical reading. The threshold has TWO determinants that should not be
conflated. The SUBSTANTIVE margin (how much standardized imbalance you are willing
to call "balanced") is an effect-size judgment, NOT a function of sample size --
0.25 SD is the conventional margin in the matching literature (Rubin 2001 /
Stuart 2010; see the provenance section below). The STATISTICAL floor above is a
function of ESS, $k$, and matching
tightness. Use the larger of the two. For typical designs (ESS in the hundreds,
$k \sim 10$) the floor is ~0.05-0.15, comfortably below 0.25, so the substantive
margin governs and a fixed 0.25 is fine. The floor only BINDS -- and 0.25 becomes
unreliable -- in the small-ESS / many-covariate corner (e.g. ESS=25, $k=20$ above),
where the data cannot certify balance at the 0.25 margin. This is the
equivalence-testing answer: fix the substantive margin, and whether you can
CONCLUDE balance at it depends on ESS (power) and $k$ (multiplicity)
(Hartman & Hidalgo 2018). A sensible package default: flag at $|\mathrm{SMD}| \ge
0.25$, and warn when the floor for the design's ESS and $k$ approaches 0.25 (the
screen cannot then reliably distinguish 0.25-imbalance from noise).

## B9 -- the size cost of pre-filtering (why decision D is "keep separate")

The magnitude screen must be read-only: it must NOT choose which covariates the
omnibus is computed on (decision D). This experiment shows the size cost of
violating that. On NULL data (no systematic imbalance), select the covariates with
$|\mathrm{SMD}| \ge \tau$ and run the omnibus on ONLY those; measure how often it
rejects. 200 pairs, $k = 10$ and $20$ (so $\mathrm{rank}(V_d) = k$, far below the
residual df of 200 -- no degeneracy). `cond.size` = rejection rate among the runs
where at least one covariate was selected; `overall.size` counts no-selection as
no-rejection.

```{r B9}
b9_nrep <- 2000L; b9_taus <- c(0.10, 0.15, 0.20, 0.25)
set.seed(20260620)
b9 <- do.call(rbind, lapply(c(10, 20), function(k) {
  base <- numeric(b9_nrep)
  psel <- matrix(NA_real_, b9_nrep, length(b9_taus))
  for (r in seq_len(b9_nrep)) {
    df <- make_pairs(200, rep(1, k), rep(0, k), rho = 0.5)   # NULL: mu = 0
    cn <- covariate_names(df)
    smds <- vapply(cn, function(c0) smd(df, c0), numeric(1))
    base[r] <- run_omnibus(df)$p_d2
    for (j in seq_along(b9_taus)) {
      sel <- cn[abs(smds) >= b9_taus[j]]
      if (length(sel)) psel[r, j] <- run_omnibus(df, covs = sel)$p_d2
    }
  }
  data.frame(k = k, all_cov_size = round(mean(base <= 0.05), 3), tau = b9_taus,
             P_select = round(colMeans(!is.na(psel)), 3),
             cond_size = round(vapply(seq_along(b9_taus),
                 function(j) mean(psel[!is.na(psel[, j]), j] <= 0.05), numeric(1)), 3),
             overall_size = round(colSums(psel <= 0.05, na.rm = TRUE) / b9_nrep, 4))
}))
b9
```

What it means. The all-covariate omnibus holds its size (0.049 at $k=10$; 0.036 at
$k=20$, slightly conservative). But CONDITIONAL on selection the pre-filtered
omnibus rejects 0.98--1.00 of the time once $\tau \ge 0.15$ -- it essentially
always rejects when it fires -- and 0.2--0.26 even at $\tau=0.10$ (4--5x nominal).
The p-value reported on the selected covariates is invalid. The OVERALL rate is
worst at lower thresholds ($\tau = 0.10$--$0.15$: ~0.12--0.19, several times
nominal, because selection fires often) and looks "safe" at $\tau=0.25$ (~0.001)
ONLY because selection rarely fires at 200 pairs -- not because the test is valid
(its conditional size is 1.0). At high $\tau$ the `cond.size` cell rests on few
firings (P(select) is tiny), so read it as "degenerate, always rejects," not as a
precise number. Bottom line: pre-filtering the omnibus by observed imbalance is
invalid; decision D = keep the magnitude screen out of the omnibus computation.
(Note the medical-tradition 0.10 threshold inflates the overall rate the most --
another reason the threshold choice matters; see Provenance below.)

## Provenance of the 0.25-SD rule

A literature check (2026-06-17) corrected a misattribution we had been repeating.
We had called the 0.25 cutoff the "Rosenbaum-Rubin rule." That is wrong in two
ways, and the threshold and denominator are not as settled as a single "0.25"
suggests.

- The "0.25" in **Rosenbaum & Rubin (1985, *The American Statistician* 39:33-38)**
  is a CALIPER width on the linear propensity score (a matching tolerance), NOT a
  balance threshold. Stuart (2010) quotes them: "Rosenbaum and Rubin (1985b)
  generally suggest a caliper of 0.25 standard deviations of the linear propensity
  score." Citing R&R 1985 for a balance cutoff is a misattribution.
- The 0.25 BALANCE threshold comes from **Rubin (2001, *Health Services & Outcomes
  Research Methodology* 2:169-188)** as restated by **Stuart (2010, *Statistical
  Science* 25:1-21)**: "the absolute standardized differences of means should be
  less than 0.25 ... (Rubin, 2001)." Rubin's own prose does not say 0.25 -- his
  Condition 1 is "less than half a standard deviation," and for the PROPENSITY
  SCORE mean difference, not individual covariates. So the flat "0.25" is Stuart's
  compression.
- TWO CAMPS ON THE NUMBER. Matching / political-methodology tradition (Rubin 2001;
  Stuart 2010; Ho-Imai-King-Stuart 2007 / MatchIt) uses **0.25**. Pharmaco-
  epidemiology / medical tradition (Normand et al. 2001, *J Clin Epi* 54:387-398;
  Austin 2009, *Stat Med* 28:3083-3107; Austin 2011, *Multivar Behav Res*
  46:399-424) uses **0.1** ("a standardized difference less than 0.1 ... indicates
  a negligible difference," Austin 2011). Cohen's d "small" = 0.2 is a separate
  lineage. The 0.1 vs 0.25 split is a real, deliberate disagreement.
- TWO CAMPS ON THE DENOMINATOR. Stuart / MatchIt use the full TREATED-group SD
  fixed at pre-matching; Austin / medical use the POOLED SD
  $\sqrt{(s_t^2 + s_c^2)/2}$. **RItools' `std.diff` uses the pooled SD** -- the
  Austin convention, which traditionally pairs with 0.1, not 0.25. So a 0.25 cutoff
  on RItools' pooled-SD `std.diff` mixes conventions; the default value AND
  denominator deserve an explicit, documented choice (the `covariate.scales`
  argument can fix the denominator to pre-matching SDs if we want the Stuart
  convention).
- PRIOR ART FOR OUR CENTRAL STANCE. **Ho, Imai, King & Stuart (2007, *Political
  Analysis* 15:199-236)** argue balance is an in-sample property and that
  hypothesis tests / p-values should NOT be used as balance criteria, because
  rejection depends on sample size (which matching changes). That is exactly the
  large-sample point in A3, and it supports presenting the omnibus as a direction
  test while leading with the magnitude (`std.diff`).

Confidence: verbatim quotes were confirmed from the primary text of Rubin (2001),
Stuart (2010), Austin (2009, 2011). The R&R (1985) caliper wording and Normand et
al. (2001) are from reliable secondary quotation (originals paywalled); confirm
those two primaries directly before citing them in a paper.

# Conclusions / package implications

1. **The collapse is real, exact, and unescapable** by any statistic on the
   within-set permutation reference -- d^2 and ACAT alike (A1, B3), and robust to
   correlation. The omnibus reads the systematic tilt relative to the within-pair
   noise, scaled by sample size; it does not read the size of imbalance, and in
   large samples rejects on trivial bias (A3). For users: report a fixed-scale
   magnitude or equivalence result as the primary balance summary, and present the
   omnibus as a test of systematic direction, not a balance-quality score.

2. **Two screens, two jobs -- keep them separate.**
   (a) *Design screen* (drop exactly-matched directions): build it for honest
   abstention plus numerical safety (B1b), a modest power gain (B1), size-preserving
   (B2), ungameable (B5) -- but NOT a collapse fix (B3). It must work on
   **directions** (singular values of $V_d$), not raw covariates, because under
   correlation the degenerate direction is a linear combination $w_j$ misses (B6).
   Implementation: a relative tolerance at `R/Design.R:991` on the eigenvalues of
   $V_d$ relative to the complete-randomization covariance, reported as a profile
   over `tau`, with graceful degradation replacing the `stop()` at `R/utils.R:381`.
   (b) *Magnitude screen* (the Love-plot / 0.25-SD rule): this is the one that
   tracks balance and escapes the collapse (B7). It is the primary balance
   **report**, not a test pre-filter (assignment-dependent; would inflate the
   omnibus if fed to it). The package should present per-covariate SMDs against a
   0.25 reference alongside, not inside, the omnibus.

3. **Keep d^2 AND switch to ACAT when needed (B4)** -- stronger under correlation.
   ACAT wins the dense case once covariates are correlated and is the only one with
   power in high dimension (where d^2 freezes near 0.45). Concrete trigger: switch
   when #covariates (after screening) $\ge \mathrm{rank}(V_d)$; consider preferring
   ACAT or reporting both more broadly given the dense-case flip. Neither omnibus
   escapes the collapse, so this is about power and numerical robustness, not about
   seeing the size of imbalance.

# Open items

- Vary the correlation: sweep `rho` (and try AR(1)) to map how the B4 dense flip
  and the d^2/ACAT crossover depend on correlation strength and structure.
- Production direction screen: derive eigenvalues from the engine's `tcov` plus a
  complete-randomization reference, not the base-R within-vs-total proxy.
- Equivalence framing for the magnitude screen: choose the default margin (0.25 SD
  in the matching tradition vs 0.1 in the medical tradition -- see provenance
  section) and whether to report a formal equivalence test (Hartman & Hidalgo 2018)
  or just SMDs vs the reference line. Also decide the denominator (RItools uses
  pooled SD; the 0.25 lineage uses the treated-group SD).
- If we proceed to package code: write the testthat tests FIRST (per
  CLAUDE_CODING.md) -- these experiments are the statistical-principle tests --
  then implement the screen, graceful degradation, and the ACAT switch, then
  `devtools::check()`.

```{r session}
#| echo: false
sessionInfo()
```
